---
title: "Recovering a High-Dimensional Vine Copula with hdcd"
subtitle: "End-to-end validation: simulate a known mixed-family vine copula, recover it, and check the results"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 9)
```

# Purpose

This notebook is an end-to-end validation exercise for the `hdcd` library
(this repository): we

1. simulate data from a **known, high-dimensional, mixed-family vine
   copula** (Clayton, Gumbel, Frank, Gaussian, and Student-*t* pair-copulas,
   with non-uniform, heterogeneous marginals on top),
2. **recover it** with `hdcd`'s full pipeline (marginal smoothing → copula
   transform → distance-correlation dependence matrix → persistent-topology
   ordering → simulated-annealing DAG search → fixed-DAG fit),
3. time every stage, and
4. check the result three ways: does the recovered DAG match the true
   dependency structure, does its fitted score compare well to fitting the
   *true* DAG directly, and do the fitted conditional copula densities
   actually match the known, closed-form true densities?

We report timings throughout and finish with conditional density plots
comparing hdcd's nonparametric fit against ground truth.

# Setup

```{r libraries}
# This notebook lives at <repo_root>/notebooks/vine_copula_recovery.Rmd.
this_file <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NA_character_)
if (is.na(this_file) || !nzchar(this_file)) this_file <- file.path(getwd(), "vine_copula_recovery.Rmd")
repo_root <- normalizePath(file.path(dirname(this_file), ".."), mustWork = FALSE)
if (!file.exists(file.path(repo_root, "r", "DESCRIPTION"))) {
  # Fallback: assume the working directory is already the repo root.
  repo_root <- normalizePath(getwd(), mustWork = FALSE)
}
stopifnot(file.exists(file.path(repo_root, "r", "DESCRIPTION")))

if (!requireNamespace("hdcd", quietly = TRUE)) {
  message("Installing hdcd from source (", file.path(repo_root, "r"), ") ...")
  install.packages(file.path(repo_root, "r"), repos = NULL, type = "source")
}

suppressMessages({
  library(hdcd)
  library(copula)
  library(ggplot2)
  library(reshape2)
  library(knitr)
})

set.seed(20260819)
```

# A known, high-dimensional, mixed-family vine copula

## Construction

We build the ground truth as a **first-order (D-vine, tree-1) Markov-chain
copula**: variables $1, \ldots, d$ in a fixed order, where $U_{i+1}$ is
generated from a bivariate pair-copula conditional on $U_i$ alone. This is
a genuine, valid vine copula (a truncated D-vine) — the simplest member of
the D-vine family — and by mixing *different* pair-copula families and
strengths across the chain we still get a rich, high-dimensional,
non-Gaussian dependency structure.

**Why not a full multi-tree R-vine?** We tried one first. Chaining
`copula::cCopula()`'s conditional-inverse (h-function) calls through a
second tree hit a genuine numerical fragility: `uniroot()` inside the
package's Archimedean inverse-Rosenblatt implementation would occasionally
fail (`f.lower = f(lower) is NA`) once an intermediate conditional value
landed very close to the $[0,1]$ boundary after two chained inversions.
Rather than ship a demo whose *ground truth* might be silently wrong, we
kept to the tree-1 construction, which we verify below is exactly correct
(uniform marginals, matching Kendall's $\tau$), and mix five different
copula families across it for genuine complexity. Even a single-tree chain
directly exercises the part of `hdcd` we most want to validate: sparse DAG
recovery plus nonparametric conditional-copula shape estimation.

```{r vine-construction}
d <- 10L
n <- 2000L
eps <- 1e-6 # keeps chained h-function inversions off the exact [0,1] boundary

# One pair-copula per edge (i, i+1), i = 1..d-1: a deliberately varied mix.
families <- list(
  claytonCopula(3.0),   # 1 -> 2   strong lower-tail dependence
  gumbelCopula(2.2),    # 2 -> 3   upper-tail dependence
  frankCopula(5.0),     # 3 -> 4   symmetric, no tail dependence
  normalCopula(0.65),   # 4 -> 5   Gaussian
  tCopula(0.5, df = 4), # 5 -> 6   symmetric tail dependence (Student-t)
  claytonCopula(1.8),   # 6 -> 7   weaker lower-tail dependence
  gumbelCopula(3.0),    # 7 -> 8   strong upper-tail dependence
  frankCopula(3.5),     # 8 -> 9   symmetric, weaker
  normalCopula(0.45)    # 9 -> 10  Gaussian, weaker
)
family_names <- c("Clayton(3.0)", "Gumbel(2.2)", "Frank(5.0)", "Gaussian(0.65)",
                   "t(0.5, df=4)", "Clayton(1.8)", "Gumbel(3.0)", "Frank(3.5)", "Gaussian(0.45)")
stopifnot(length(families) == d - 1)

hinv <- function(u_given, w, cop) cCopula(cbind(u_given, w), copula = cop, indices = 2, inverse = TRUE)

w <- matrix(runif(n * d), ncol = d)
w <- pmin(pmax(w, eps), 1 - eps)

U <- matrix(NA_real_, n, d)
U[, 1] <- w[, 1]
for (i in 2:d) {
  u_prev <- pmin(pmax(U[, i - 1], eps), 1 - eps)
  U[, i] <- hinv(u_prev, w[, i], families[[i - 1]])
  U[, i] <- pmin(pmax(U[, i], eps), 1 - eps)
}
```

## Sanity check: is the ground truth actually correct?

A valid copula sample must have exactly Uniform(0,1) margins, and the
Kendall's $\tau$ between adjacent variables should match each pair-copula's
known $\tau$. If either check failed, nothing downstream would be worth
trusting.

```{r vine-sanity-check}
ks_pvalues <- sapply(1:d, function(j) ks.test(U[, j], "punif")$p.value)
empirical_tau <- sapply(1:(d - 1), function(i) cor(U[, i], U[, i + 1], method = "kendall"))
true_tau <- sapply(families, tau)

sanity <- data.frame(
  edge = paste0(1:(d - 1), " -> ", 2:d),
  family = family_names,
  true_kendall_tau = round(true_tau, 3),
  empirical_kendall_tau = round(empirical_tau, 3),
  marginal_KS_pvalue_ui = round(ks_pvalues[-d], 3)
)
kable(sanity, caption = "Ground-truth construction check: empirical vs. true Kendall's tau, per edge")
```

All Kolmogorov–Smirnov p-values should be comfortably large (no evidence
against uniformity), and the empirical Kendall's $\tau$ values should sit
close to their true values. This is our guarantee that the "true" copula
we're about to try to recover is exactly what we think it is.

## From copula scale to realistic, non-uniform data

`hdcd` is meant to work on **raw data**, not pre-transformed copula
coordinates — it fits its own marginals. So we push $U$ through a
heterogeneous set of marginal distributions (Normal, Exponential, and
several Normals with different means) to get a realistic-looking raw
dataset $X$, throwing away direct access to $U$ except for validation.

```{r apply-marginals}
X <- U
X[, 1] <- qnorm(U[, 1], mean = 0, sd = 2)
X[, 2] <- qexp(U[, 2], rate = 1)
for (j in 3:d) X[, j] <- qnorm(U[, j], mean = j, sd = 1)

colnames(X) <- paste0("X", 1:d)
kable(t(apply(X, 2, function(col) c(min = min(col), mean = mean(col), max = max(col)))),
      caption = "Raw data X: per-column summary (heterogeneous marginals)")
```

## What the pairwise dependence actually looks like

A quick look at three adjacent pairs, chosen to show three different
pair-copula shapes: Clayton (lower-tail clustering), Gumbel (upper-tail
clustering), and Gaussian (symmetric, elliptical).

```{r pairwise-scatter, fig.height=3.2}
pairs_df <- rbind(
  data.frame(u = U[, 1], v = U[, 2], edge = "1->2  Clayton(3.0)"),
  data.frame(u = U[, 2], v = U[, 3], edge = "2->3  Gumbel(2.2)"),
  data.frame(u = U[, 4], v = U[, 5], edge = "4->5  Gaussian(0.65)")
)
ggplot(pairs_df, aes(u, v)) +
  geom_point(alpha = 0.25, size = 0.6) +
  facet_wrap(~edge) +
  coord_equal() +
  theme_minimal() +
  labs(title = "Copula-scale pairwise dependence for three representative edges",
       x = expression(U[i]), y = expression(U[i+1]))
```

# Fitting hdcd

We now forget we ever saw `U` or the true DAG, and hand `hdcd` only the raw
data `X`. We run the pipeline stage by stage (rather than calling the
single `hdcd_fit()` wrapper) purely so we can time each stage separately;
`hdcd_fit()` performs exactly the same steps internally.

```{r fit-stages}
timings <- list()
tic <- function() Sys.time()
toc <- function(t0, label) {
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
  timings[[label]] <<- elapsed
  invisible(elapsed)
}

# ---- Stage 1: marginal fitting (spec section 2) ----
t0 <- tic()
marginals <- lapply(1:d, function(j) hdcd:::.new_marginal(X[, j]))
toc(t0, "1. Marginal fitting (d columns)")

# ---- Stage 2: copula transform (spec section 4) ----
t0 <- tic()
U_hat <- matrix(NA_real_, n, d)
for (j in 1:d) U_hat[, j] <- hdcd:::.transform_to_copula(marginals[[j]], X[, j])
toc(t0, "2. Copula transform (d columns)")

# ---- Stage 3: dependence matrix (spec section 5) ----
t0 <- tic()
dm_ext <- hdcd:::.compute_dependence_matrix(U_hat, n, d)
toc(t0, "3. Distance-correlation dependence matrix")

# ---- Stage 4: persistent-topology ordering (spec section 6) ----
t0 <- tic()
topo_ext <- hdcd:::.compute_topology(dm_ext)
ordering <- hdcd:::.topology_ordering(topo_ext)
toc(t0, "4. Persistent-topology ordering")

# ---- Stage 5: simulated-annealing DAG search (spec sections 7, 17) ----
max_parents <- 2L
lambda_edge <- 0.05
bernstein_degree <- 4L
lambda_roughness <- 0.15
holdout_fraction <- 0.25
seed <- 1L
local_seed <- seed + 1L

t0 <- tic()
annealed <- hdcd:::.run_annealing_c(
  U_hat, n, d, ordering, max_parents, lambda_edge,
  bernstein_degree, lambda_roughness, holdout_fraction, local_seed,
  initial_temperature = 0.5, cooling_rate = 0.95,
  max_iterations = 600L, restarts = 3L,
  p_add = 1.0, p_remove = 1.0, p_swap = 1.0, anneal_seed = seed
)
toc(t0, "5. Simulated-annealing DAG search (1800 total proposals)")

# ---- Stage 6: final fixed-DAG fit (spec sections 14, 28) ----
t0 <- tic()
dag_fit_ext <- hdcd:::.dag_fit_c(U_hat, n, d, annealed$dag, bernstein_degree,
                                  lambda_roughness, holdout_fraction, local_seed)
toc(t0, "6. Final fixed-DAG fit")

timing_df <- data.frame(
  stage = names(timings),
  seconds = round(unlist(timings), 3)
)
timing_df <- rbind(timing_df, data.frame(stage = "TOTAL", seconds = round(sum(unlist(timings)), 3)))
kable(timing_df, caption = sprintf("Stage-by-stage timing (d=%d, n=%d)", d, n))
```

```{r assemble-model}
# Assemble the same hdcd_model structure hdcd_fit() would return, so the
# rest of the notebook can use the normal public API.
model <- list(
  marginals = marginals, dag = annealed$dag, dag_fit = dag_fit_ext,
  dependence_matrix_ext = dm_ext, topology_ext = topo_ext, d = d,
  max_parents = max_parents, bernstein_degree = bernstein_degree,
  lambda_roughness = lambda_roughness, holdout_fraction = holdout_fraction,
  local_seed = local_seed, best_score = annealed$score,
  score_trace = annealed$score_trace, accepted_trace = annealed$accepted_trace,
  acceptance_rate = annealed$acceptance_rate, U = U_hat, X = X
)
class(model) <- "hdcd_model"
print(model)
```

# Did it recover the structure?

## Recovered vs. true DAG

```{r dag-comparison}
recovered_edges <- hdcd_dag(model)
true_edges <- cbind(1:(d - 1), 2:d)

edge_key <- function(e) paste(pmin(e[, 1], e[, 2]), pmax(e[, 1], e[, 2]), sep = "-")
recovered_skeleton <- edge_key(recovered_edges)
true_skeleton <- edge_key(true_edges)

comparison <- data.frame(
  true_edge = paste0(true_edges[, 1], " -> ", true_edges[, 2]),
  recovered_as = sapply(seq_len(nrow(true_edges)), function(i) {
    key <- true_skeleton[i]
    idx <- which(recovered_skeleton == key)
    if (length(idx) == 0) return("MISSED")
    re <- recovered_edges[idx, ]
    if (re[1] == true_edges[i, 1]) "exact match" else sprintf("reversed (%d -> %d)", re[1], re[2])
  })
)
kable(comparison, caption = "Every true edge: was it recovered, and in the correct direction?")

n_exact <- sum(comparison$recovered_as == "exact match")
n_reversed <- sum(grepl("reversed", comparison$recovered_as))
n_missed <- sum(comparison$recovered_as == "MISSED")
n_spurious <- nrow(recovered_edges) - (n_exact + n_reversed)

cat(sprintf(
  "Skeleton edges recovered: %d / %d (%d exact direction, %d reversed). Spurious extra edges: %d.\n",
  n_exact + n_reversed, nrow(true_edges), n_exact, n_reversed, n_spurious
))
```

A reversed edge is not a bug: `hdcd`'s reference DAG is an optimized
**statistical factorization for density estimation**, not a causal claim
(this is stated explicitly throughout the library's own documentation and
tested directly in Milestone 9). For a chain, the pairwise dependence
between $U_i$ and $U_{i+1}$ is symmetric regardless of which one is
labelled "parent," so a direction flip on one edge — while everything
else matches — is exactly the kind of statistically-near-equivalent
alternative factorization the annealing search is allowed to land on.

## Dependence matrix and persistent-topology ordering

```{r dependence-heatmap, fig.height=5}
dm <- hdcd_dependence_matrix(model)
dm_long <- melt(dm, varnames = c("row", "col"), value.name = "dCor")
ggplot(dm_long, aes(col, row, fill = dCor)) +
  geom_tile() +
  scale_fill_viridis_c(limits = c(0, 1)) +
  scale_y_reverse() +
  coord_equal() +
  theme_minimal() +
  labs(title = "Fitted pairwise distance-correlation matrix", x = "column", y = "column")
```

The chain structure is clearly visible as a bright band along the
super-/sub-diagonal (adjacent variables are strongly dependent), fading
away from it — exactly the pattern a first-order Markov chain should
produce.

```{r ordering}
cat("Persistent-topology ordering:", hdcd_ordering(model), "\n")
cat("(true generative order was 1, 2, ..., 10 -- any ordering consistent\n",
    " with the chain's actual dependency strengths is a valid recovery.)\n")
```

## Did the annealing search actually converge?

```{r annealing-trace, fig.height=3.5}
trace_df <- data.frame(iteration = seq_along(model$score_trace), score = model$score_trace)
ggplot(trace_df, aes(iteration, score)) +
  geom_line(alpha = 0.5) +
  geom_hline(yintercept = model$best_score, color = "firebrick", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Simulated-annealing score trace (J(G) after each iteration)",
       subtitle = sprintf("acceptance rate = %.2f; dashed line = best score found (%.4f)",
                           model$acceptance_rate, model$best_score),
       x = "iteration (across all restarts)", y = "J(G)  [lower is better]")
```

$J(G)$ should trend down and then flatten out as the search converges on a
good structure — a rising-then-flat noisy trace, not one that's still
improving steeply at the end, is what confirms the iteration budget was
sufficient.

# How good is the recovered DAG's fit, really?

Structural mismatches on individual edges don't necessarily mean a worse
*fit*. We can check directly: fit the true DAG explicitly with
`hdcd_fit_dag()` (over the exact same training data) and compare held-out
KL scores with `hdcd_score_dag()` (spec section 19).

```{r true-dag-fit}
t0 <- tic()
true_fit <- hdcd_fit_dag(model, true_edges)
true_fit_time <- toc(t0, "True-DAG fit_dag() (for comparison only)")

delta_kl <- hdcd_score_dag(model, true_fit)
cat(sprintf("fit_dag(true DAG) took %.2f seconds\n", true_fit_time))
cat(sprintf(
  "Delta_KL(true DAG vs. recovered/annealed DAG) = %.4f\n(negative means the TRUE dag scores BETTER; near 0 means the annealed DAG is nearly as good)\n",
  delta_kl
))
```

# Do the fitted conditional densities actually match?

This is the most direct test of `hdcd`'s core nonparametric machinery
(the centered Bernstein tensor kernel plus Sinkhorn normalization,
Milestones 5-6): for every true edge, does `hdcd`'s fitted conditional
copula density $\hat c(u \mid z)$ — evaluated from the **true-DAG fit**,
so this is purely about density-shape accuracy, independent of the
discrete structure-search question above — match the known, closed-form
conditional density of the true pair-copula?

```{r conditional-density-grid, fig.height=14}
u_grid <- seq(0.02, 0.98, length.out = 60)
q_values <- c(0.1, 0.5, 0.9)

cond_rows <- list()
for (i in seq_len(d - 1)) {
  node <- i + 1L # child column (1-indexed), matches true_fit's node numbering
  for (q in q_values) {
    fitted <- exp(hdcd:::.local_fit_conditional_log_density(true_fit$dag_fit, node, u_grid, q))
    true_dens <- dCopula(cbind(rep(q, length(u_grid)), u_grid), families[[i]])
    cond_rows[[length(cond_rows) + 1]] <- data.frame(
      edge = sprintf("%d: %d->%d  %s", i, i, i + 1, family_names[i]),
      q = factor(sprintf("z = %.1f", q), levels = sprintf("z = %.1f", q_values)),
      u = u_grid,
      density = c(fitted, true_dens),
      source = rep(c("hdcd (fitted)", "true"), each = length(u_grid))
    )
  }
}
cond_df <- do.call(rbind, cond_rows)

ggplot(cond_df, aes(u, density, color = source, linetype = source)) +
  geom_line(linewidth = 0.7) +
  facet_grid(edge ~ q, scales = "free_y") +
  scale_color_manual(values = c("hdcd (fitted)" = "steelblue", "true" = "black")) +
  theme_minimal(base_size = 9) +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), legend.position = "top") +
  labs(title = "Fitted vs. true conditional copula density c(u | z) for every true edge",
       x = expression(u[child]), y = "density")
```

## Summary agreement metrics

```{r conditional-density-summary}
agreement <- do.call(rbind, lapply(seq_len(d - 1), function(i) {
  node <- i + 1L
  per_q <- sapply(q_values, function(q) {
    fitted <- exp(hdcd:::.local_fit_conditional_log_density(true_fit$dag_fit, node, u_grid, q))
    true_dens <- dCopula(cbind(rep(q, length(u_grid)), u_grid), families[[i]])
    c(cor = cor(fitted, true_dens), rmse = sqrt(mean((fitted - true_dens)^2)))
  })
  data.frame(
    edge = sprintf("%d->%d", i, i + 1), family = family_names[i],
    mean_correlation = round(mean(per_q["cor", ]), 3),
    mean_rmse = round(mean(per_q["rmse", ]), 3)
  )
}))
kable(agreement, caption = "Per-edge agreement between hdcd's fitted density and the true density, averaged over z in {0.1, 0.5, 0.9}")

cat(sprintf("\nOverall mean correlation across all edges/conditioning values: %.3f\n",
            mean(agreement$mean_correlation)))
```

The per-edge table above is not uniform: the two smooth, no-corner-singularity
families (Frank, Gaussian) sit at correlation $\geq 0.97$, while the three
edges built from **tail-dependent** families (Clayton at edges 1 and 6,
Gumbel at edges 2 and 7) sit noticeably lower (0.90-0.96). That is not
random scatter — it is a real, systematic shape mismatch, worth
understanding rather than averaging away into the single headline number
above.

## Known limitation: sharp tail-dependence corners are under-fit

Look closely at the row-1 and row-6 panels (Clayton) at $z=0.1$, and the
row-7 panel (Gumbel) at $z=0.9$, in the density grid above: the **true**
density has a sharp, narrow spike right at the copula corner (Clayton's
lower-tail dependence concentrates mass as $u, z \to 0$; Gumbel's
upper-tail dependence does the same as $u, z \to 1$), while `hdcd`'s
**fitted** curve is visibly flatter and shorter there, even though it
tracks the true curve well everywhere else. This is the actual thing that
prompted a closer look at "densities that don't line up."

The first hypothesis was that a single, hand-picked, *global*
`lambda_roughness = 0.15` — applied identically to all nine edges — was
over-smoothing specifically these sharply-peaked edges, and that a
lighter, per-edge penalty would fit them better. `hdcd` now supports
exactly that as an option (`lambda_roughness_grid`, see `DECISIONS.md`):
a small validation grid, searched independently per node via an *inner*
train/validation split that never touches the outer holdout, entirely
outside the annealing search (spec section 18). Applying it directly
tests the hypothesis rather than just asserting it.

```{r roughness-grid-fit}
roughness_grid <- c(0.02, 0.05, 0.1, 0.15, 0.3, 0.6, 1.0)
t0 <- tic()
auto_fit <- hdcd:::.dag_fit_c(model$U, n, d, true_fit$dag, bernstein_degree, lambda_roughness,
                               holdout_fraction, local_seed,
                               lambda_roughness_grid = roughness_grid,
                               roughness_validation_fraction = 0.3)
auto_fit_time <- toc(t0, "Per-node auto lambda_roughness fit_dag() (diagnostic only)")

selected <- do.call(rbind, lapply(seq_len(d - 1), function(i) {
  node <- i + 1L
  data.frame(
    edge = sprintf("%d->%d", i, i + 1), family = family_names[i],
    selected_lambda = hdcd:::.local_fit_selected_lambda_roughness(auto_fit, node),
    holdout_score_auto = round(hdcd:::.dag_fit_holdout_scores(auto_fit, d)[node], 4),
    holdout_score_fixed_0.15 = round(hdcd:::.dag_fit_holdout_scores(true_fit$dag_fit, d)[node], 4)
  )
}))
kable(selected, caption = sprintf(
  "Per-node lambda_roughness selected from {%s} by inner-validation, vs. this notebook's fixed 0.15",
  paste(roughness_grid, collapse = ", ")
))
cat(sprintf("Delta_KL(fixed@0.15 vs. auto-selected) = %.4f\n",
            hdcd:::.dag_fit_kl_difference(true_fit$dag_fit, auto_fit)))
```

**The hypothesis does not survive contact with the data.** Every edge —
Clayton and Gumbel included — selects a lambda at or above this
notebook's original 0.15, not below it; several land higher (0.3). A
finer, continuous sweep (fixed `lambda_roughness`, no grid search)
confirms this is a genuine interior optimum, not an artifact of the
grid's spacing:

```{r roughness-fine-sweep, fig.height=3.2}
fine_lambdas <- c(0.01, 0.02, 0.05, 0.1, 0.15, 0.3, 0.6, 1.0, 2.0)
fine_sweep <- do.call(rbind, lapply(c(edge1 = 2L, edge7 = 8L, edge3 = 4L), function(node) {
  do.call(rbind, lapply(fine_lambdas, function(lr) {
    fit <- hdcd:::.dag_fit_c(model$U, n, d, true_fit$dag, bernstein_degree, lr, holdout_fraction, local_seed)
    data.frame(edge = c(`2` = "1->2 Clayton(3.0)", `8` = "7->8 Gumbel(3.0)", `4` = "3->4 Frank(5.0)")[[as.character(node)]],
               lambda_roughness = lr,
               holdout_score = hdcd:::.dag_fit_holdout_scores(fit, d)[node])
  }))
}))
ggplot(fine_sweep, aes(lambda_roughness, holdout_score)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = 0.15, linetype = "dashed", color = "firebrick") +
  scale_x_log10() +
  facet_wrap(~edge, scales = "free_y") +
  theme_minimal() +
  labs(title = "Held-out score vs. lambda_roughness (log scale), fixed bernstein_degree",
       subtitle = "dashed line = this notebook's fixed 0.15",
       x = "lambda_roughness (log scale)", y = "held-out log-density score")
```

The same interior-optimum shape shows up for the Clayton edge, the
Gumbel edge, and even the smooth Frank edge (which has no corner at
all) — none of them want much lighter regularization than 0.15, and all
of them clearly punish going much lighter. `Delta_KL` in the table above
is slightly *positive* for the auto-selected fit relative to the fixed
default: per-node selection pays a real cost here (each candidate is fit
on a smaller *inner*-train subset, which is why it exists to begin with
— see `DECISIONS.md`), and on this dataset that cost isn't repaid,
because 0.15 was already close to optimal everywhere it was tested.

**Revised conclusion.** The original diagnosis — eyeballing the fitted
curve against the true curve and concluding "a lighter penalty would fit
this corner better" — was measuring agreement with a density that is only
knowable here because this is a synthetic exercise with ground truth.
*Held-out log-likelihood*, the criterion `hdcd` (and any real, non-
synthetic application) actually optimizes, disagrees: it says 0.15 is
already a good choice for these edges, lighter regularization measurably
overfits, and the visual corner-flattening is not primarily a smoothing
problem. What's left is more consistent with a basis-expressivity
limit — a degree-4 centered Bernstein tensor is a smooth, globally
bounded polynomial surface, and no amount of retuning `lambda_roughness`
lets a bounded polynomial reproduce an unbounded corner singularity. This
is a more honest, if less tidy, takeaway than the first pass reached, and
exactly what building the feature properly (rather than assuming the fix)
was for.

## Following up: does raising bernstein_degree, done properly, help?

The natural next lever, and a genuinely different question from the
`lambda_roughness` one above: is a degree-4 centered Bernstein tensor
just too coarse a *basis* to represent a sharp corner, independent of how
it's regularized? `hdcd` now supports raising `bernstein_degree`
per-node too (`bernstein_degree_grid`), gated by an empirical
tail-dependence coefficient (a standard nonparametric EVT/copula
estimator — see `DECISIONS.md`'s "tail-dependence-informed
`bernstein_degree` selection") so the extra flexibility (and its cost) is
only spent on edges that actually show tail dependence, and searched
*jointly* with `lambda_roughness` this time — the earlier, very first
degree check in this notebook held lambda fixed while raising degree,
which confounded "does more basis flexibility help" with "is it
regularized correctly for its new size."

```{r degree-grid-fit}
degree_grid <- c(4L, 6L, 8L, 10L)
lambda_grid_2 <- c(0.05, 0.1, 0.15, 0.3)
t0 <- tic()
degree_fit <- hdcd:::.dag_fit_c(model$U, n, d, true_fit$dag, bernstein_degree, lambda_roughness,
                                 holdout_fraction, local_seed,
                                 lambda_roughness_grid = lambda_grid_2, roughness_validation_fraction = 0.3,
                                 bernstein_degree_grid = degree_grid, tail_dependence_gate = 0.05)
degree_fit_time <- toc(t0, "Tail-dependence-gated joint (degree, lambda) fit_dag() (diagnostic only)")

degree_summary <- do.call(rbind, lapply(seq_len(d - 1), function(i) {
  node <- i + 1L
  q <- if (i %in% c(2, 7)) 0.9 else 0.1 # Gumbel edges: upper tail; else lower
  fitted_fixed <- exp(hdcd:::.local_fit_conditional_log_density(true_fit$dag_fit, node, u_grid, q))
  fitted_auto <- exp(hdcd:::.local_fit_conditional_log_density(degree_fit, node, u_grid, q))
  true_dens <- dCopula(cbind(rep(q, length(u_grid)), u_grid), families[[i]])
  data.frame(
    edge = sprintf("%d->%d", i, i + 1), family = family_names[i],
    tail_dependence = round(hdcd:::.local_fit_max_tail_dependence(degree_fit, node), 3),
    selected_degree = hdcd:::.local_fit_selected_bernstein_degree(degree_fit, node),
    selected_lambda = hdcd:::.local_fit_selected_lambda_roughness(degree_fit, node),
    cor_degree4_fixed = round(cor(fitted_fixed, true_dens), 3),
    cor_degree_auto = round(cor(fitted_auto, true_dens), 3)
  )
}))
kable(degree_summary, caption = sprintf(
  "Tail-dependence-gated joint (degree in {%s}, lambda in {%s}) selection vs. the fixed degree=4/lambda=0.15 baseline",
  paste(degree_grid, collapse = ", "), paste(lambda_grid_2, collapse = ", ")
))
cat(sprintf("Delta_KL(fixed baseline vs. degree-auto fit) = %.4f\n",
            hdcd:::.dag_fit_kl_difference(true_fit$dag_fit, degree_fit)))
```

Two things happen at once, and they point in opposite directions:

- **The tail-dependence diagnostic itself is doing its job.** The
  measured coefficient is high for exactly the edges flagged visually
  (Clayton, Gumbel) and low for the smooth ones (Frank, Gaussian) — this
  is the same corner-concentration property being picked up
  independently, by an estimator that has never seen a fitted density
  curve.
- **Shape-correlation to the true density DOES improve on the flagged
  edges** once degree is raised jointly with a re-tuned (lighter)
  lambda — unlike the very first, unfair degree-only check earlier in
  this notebook. Both Clayton edges and both Gumbel edges gain several
  points of correlation.
- **But `Delta_KL` goes the wrong way**: the auto (higher-degree)
  fit's *held-out log-likelihood* is measurably worse than the plain
  degree=4/lambda=0.15 baseline, not better. A degree-10 tensor has
  $11 \times 11 = 121$ coefficients per edge instead of $5\times5=25$ at
  degree 4 — roughly 5x the parameters, fit from a training set this
  notebook never grew to match. It reproduces the true curve's *shape*
  more closely while generalizing *worse* to new data at $n=2000$: the
  same shape-vs-likelihood tension the `lambda_roughness` investigation
  surfaced, now showing up again one level up, in the choice of basis
  size.

**Where this leaves the original question.** Raising `bernstein_degree`
is not a free win either, at least not at this sample size: it can
literally trade held-out predictive accuracy for a prettier picture of a
density nobody could observe in a real (non-synthetic) application. Two
honest paths forward, both logged in `DECISIONS.md` rather than picked
here: grow $n$ until the extra flexibility pays for itself (this
notebook fixes $n=2000$ throughout — a genuinely separate experiment),
or reach for the heavier, more targeted fix — a true EVT tail-splice
that adds a handful of *parametric* tail-behavior parameters exactly
where the corner is, instead of dozens more general-purpose polynomial
coefficients everywhere.

## Does more data resolve the bias-variance tradeoff?

Picking up the first of those two paths directly: does growing $n$ make
the higher-`bernstein_degree` fit actually pay off in held-out
log-likelihood, not just in shape-correlation? We reran the exact same
tail-dependence-gated joint (degree, lambda) search — degree in
$\{4,6,8,10\}$, lambda in $\{0.05,0.1,0.15,0.3\}$, gate $0.05$ — against
the same true DAG, at $n \in \{2000, 4000, 8000, 16000\}$, tracking the
two most severely under-fit edges (1->2 Clayton, 7->8 Gumbel) plus the
overall (all 9 edges) `Delta_KL`.

This sweep is genuinely expensive — the joint search's cost grows worse
than linearly in $n$ (19s at $n=2000$ vs. 879s at $n=16000$ for the
9-edge search alone, on top of copula-scale data generation itself
taking up to 84s at the largest $n$), roughly half an hour in total — so
it was run once, standalone
(`notebooks/n_sweep_experiment.R`, saved to `notebooks/n_sweep_results.csv`),
rather than re-executed on every render of this notebook.

```{r n-sweep-load}
sweep_results <- read.csv("n_sweep_results.csv")
kable(sweep_results[, c("n", "edge", "family", "selected_degree", "selected_lambda",
                         "cor_fixed", "cor_auto", "delta_kl_total")],
      digits = 3, row.names = FALSE,
      caption = "Tail-dependence-gated joint search vs. sample size (one realization per n; see notebooks/n_sweep_experiment.R)")
```

```{r n-sweep-plot, fig.height=3.2}
kl_by_n <- unique(sweep_results[, c("n", "delta_kl_total")])
p1 <- ggplot(kl_by_n, aes(n, delta_kl_total)) +
  geom_line() + geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Delta_KL(fixed vs. auto), all 9 edges", subtitle = "0 = tied; positive = fixed baseline still better",
       x = "n (log scale)", y = "Delta_KL")

cor_long <- reshape2::melt(sweep_results[, c("n", "edge", "cor_fixed", "cor_auto")],
                            id.vars = c("n", "edge"), variable.name = "fit", value.name = "correlation")
p2 <- ggplot(cor_long, aes(n, correlation, color = fit, linetype = edge)) +
  geom_line() + geom_point(size = 1.5) +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Shape-correlation to true density", x = "n (log scale)", y = "correlation")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

**The first three points looked like a clean story, and then it wasn't.**
`Delta_KL` fell from 0.252 (n=2000) to 0.203 (n=4000) to 0.099 (n=8000) —
each doubling roughly halving the gap, exactly what "the tradeoff is a
data-size effect" would predict, seemingly on track to cross zero
somewhere past $n=16000$. It didn't: at $n=16000$, `Delta_KL` jumped back
up to 0.319, worse than every smaller $n$ tested, including $n=2000$.

Meanwhile the *shape*-correlation numbers stayed almost boringly stable
across the whole sweep — `cor_auto` for 1->2 sits in a narrow 0.80-0.82
band and `cor_auto` for 7->8 in 0.83-0.86, at every single $n$, always a
few points above the matching `cor_fixed`. And the search picked the
grid's most flexible option (degree 10, lambda 0.05 — literally the
edge of the search space) every single time, at every $n$, never
settling partway into the grid as more data arrived.

**Why this is a real finding, not a wasted run.** Three things are true
at once here, and they don't resolve into a tidy story:

1. The search's own selection criterion (inner-validation score) always
   wants maximum flexibility, at every sample size tested — the grid's
   boundary never stopped being attractive, which on its own is a mild
   red flag that the grid itself may be too narrow to see where the
   optimum actually sits.
2. Shape-correlation to the true density is a stable, small, consistent
   win for the higher-degree fit across the whole range tested.
3. Held-out log-likelihood — the metric that actually matters — is
   noisy enough, and moved by enough else besides raw sample size, that
   four points spanning an 8x range in $n$ produced a non-monotonic
   trend with a reversal at the *largest* $n$, not the smallest.

Point 3 matters most. **This sweep is a single realization per $n$, not
a replicated design**: each $n$ draws a fresh, independent copula sample
rather than growing one fixed dataset, so sample-to-sample variability
in exactly which points land in which train/inner-validation/holdout
split is completely confounded with the effect of $n$ itself. A
1-point-per-$n$ curve cannot distinguish "the tradeoff doesn't resolve
monotonically" from "we got an unlucky draw at $n=16000$." Properly
separating those would need several replicate seeds at every $n$ and a
confidence band — and at roughly 15 minutes of compute for a *single*
$n=16000$ replicate, getting even 5 replicates there costs over an hour,
for one edge of one synthetic experiment.

That gap — between "the question is well-posed" and "answering it
rigorously is affordable" — is itself worth sitting with. Confirming
whether more data resolves a flexibility/generalization tradeoff in a
high-dimensional, heavy-tailed dependence model is not just a modeling
question; empirically *checking* the answer runs into the same
high-dimensional, heavy-tailed cost structure the modeling problem
started with. Logged in `DECISIONS.md` as an open question rather than
resolved here: the honest result of this sweep is "inconclusive, and
expensive to make conclusive," not "yes" or "no."

## Does corner relief resolve the tradeoff, properly replicated?

The `bernstein_degree_grid` route added *more coefficients everywhere*
and paid a generalization cost for it. A cheaper, more surgical
alternative discussed as a follow-up: relax the roughness penalty
specifically near the Bernstein coefficient grid's four CORNERS — where
$u$ and $z$ are both extreme simultaneously, exactly where tail
dependence concentrates — instead of raising the coefficient count.
`hdcd` now supports this directly
(`corner_relief`, see `DECISIONS.md`'s "anisotropic (corner-relaxed)
roughness penalty"): one scalar in $[0,1)$, applied identically inside
every Theta fit, at zero extra cost per fit (unlike a grid search).

That last property — no extra fit calls — makes something the earlier
`bernstein_degree_grid` sweep couldn't afford actually affordable here:
**real replication.** Six independent seeds at every $n \in \{2000,
4000, 8000, 16000\}$, for `corner_relief` $\in \{0, 0.3, 0.6, 0.8\}$
(fixed `bernstein_degree=4`, `lambda_roughness=0.15` throughout, so
`corner_relief` is the only thing varying), on the same two worst edges
(1->2 Clayton, 7->8 Gumbel). Standalone script
`notebooks/corner_relief_sweep_experiment.R` (~48 minutes total),
results saved to `notebooks/corner_relief_sweep_results.csv` and loaded
here, not recomputed.

```{r corner-relief-sweep-load}
cr_results <- read.csv("corner_relief_sweep_results.csv")
cr_agg <- aggregate(
  cbind(cor_true, delta_kl_total) ~ n + corner_relief, data = cr_results,
  FUN = function(x) c(mean = mean(x), sd = sd(x))
)
cr_agg <- do.call(data.frame, cr_agg)
names(cr_agg) <- c("n", "corner_relief", "cor_mean", "cor_sd", "delta_kl_mean", "delta_kl_sd")
cr_agg$z_score <- ifelse(cr_agg$delta_kl_sd > 0, cr_agg$delta_kl_mean / cr_agg$delta_kl_sd, NA)
kable(cr_agg, digits = 3, row.names = FALSE,
      caption = "Mean +/- SD over 6 replicates x 2 edges (12 observations per row). z_score = Delta_KL mean / Delta_KL SD.")
```

```{r corner-relief-sweep-plot, fig.height=3.2}
p1 <- ggplot(cr_agg, aes(n, cor_mean, color = factor(corner_relief))) +
  geom_line() +
  geom_pointrange(aes(ymin = cor_mean - cor_sd / sqrt(12), ymax = cor_mean + cor_sd / sqrt(12))) +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Shape-correlation to true density", subtitle = "mean +/- SE, 12 obs/cell",
       x = "n (log scale)", y = "correlation", color = "corner_relief")

p2 <- ggplot(subset(cr_agg, corner_relief > 0), aes(n, delta_kl_mean, color = factor(corner_relief))) +
  geom_line() +
  geom_pointrange(aes(ymin = delta_kl_mean - delta_kl_sd / sqrt(12), ymax = delta_kl_mean + delta_kl_sd / sqrt(12))) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Delta_KL vs. corner_relief=0 baseline", subtitle = "mean +/- SE; 0 = tied with baseline",
       x = "n (log scale)", y = "Delta_KL", color = "corner_relief")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

**`corner_relief = 0.6` is the standout, and it holds up under
replication.** It improves shape-correlation at *every* $n$ tested (0.3
and 0.8 are more mixed — 0.8 is even slightly below baseline at
$n=2000$), and its `Delta_KL` cost, while consistently positive on
average, never reaches even 1 replicate standard deviation at any $n$
(z-scores 0.14–0.98) — squarely inside the "should not sit outside
roughly 1-2 SD above 0" bar set *before* this sweep ran (see
`DECISIONS.md`). That is a genuinely different outcome from the
`bernstein_degree_grid` sweep two sections up: there, a single
realization per $n$ swung non-monotonically across a much wider range
with no way to tell signal from noise; here, replication shows a small,
consistent shape-fit gain whose cost is not statistically distinguishable
from zero at this sample size.

One thing not to paper over: the z-score drifts upward with $n$ (0.39,
0.14, 0.86, 0.98 across the four $n$ values) — still under 1 SD
throughout, but trending toward it, not away from it. Six replicates is
enough to show the earlier degree-grid "reversal at $n=16000$" story was
mostly noise; it is not enough to rule out a genuine, slowly-growing cost
appearing at even larger $n$ than tested here. That would need either
more replicates at the high end or a purpose-built sequential design,
not attempted here.

**Bottom line:** of the three interventions tried in this notebook for
the Clayton/Gumbel corner under-fit — a validated `lambda_roughness`
(ruled out: the fixed default was already near-optimal), a validated
`bernstein_degree` (real shape gain, but a real and non-monotonic
likelihood cost that didn't resolve with more data), and now
`corner_relief` — this is the first one to clear its own pre-registered
bar. `corner_relief=0.6` is not adopted as `hdcd`'s new default here
(that is a decision for whoever maintains this library going forward,
not made unilaterally in a validation notebook), but it is the most
credible lead this investigation produced.

# Summary

```{r summary}
cat(sprintf("Dimension: d = %d, n = %d\n", d, n))
cat(sprintf("Total hdcd pipeline time: %.2f seconds\n", sum(unlist(timings))))
cat(sprintf("Skeleton edges recovered (any direction): %d / %d\n", n_exact + n_reversed, nrow(true_edges)))
cat(sprintf("Exact-direction edges: %d, reversed: %d, missed: %d, spurious: %d\n",
            n_exact, n_reversed, n_missed, n_spurious))
cat(sprintf("Delta_KL(true DAG vs. recovered DAG): %.4f\n", delta_kl))
cat(sprintf("Mean fitted-vs-true conditional density correlation: %.3f\n",
            mean(agreement$mean_correlation)))
```

**Takeaways:**

- The ground-truth generator was independently verified correct (uniform
  margins, matching Kendall's $\tau$) before being used to validate
  anything else.
- `hdcd` recovered the correct dependency *skeleton* for essentially all
  edges, in well under a minute for a 10-dimensional, 9-edge, five-family
  mixed vine copula on 2000 observations.
- Where the recovered DAG deviated from the true generative structure (a
  single reversed edge), the held-out KL score confirms it is a
  near-equally-good *statistical* factorization — precisely what the
  library's own documentation warns you to expect, and precisely why the
  reference DAG is never presented as a causal claim (spec sections 19, 34).
- The fitted conditional copula densities closely track the true,
  closed-form densities on the smooth, no-tail-dependence edges (Frank,
  Gaussian: correlation $\geq 0.97$), and reasonably well overall
  (mean correlation $\approx 0.96$), confirming the centered-
  Bernstein/Sinkhorn machinery (Milestones 5-6) is doing its job.
- The clearest weak spot: on edges with strong Archimedean tail
  dependence (Clayton, Gumbel), the fitted density visibly under-shoots
  the true density's sharp corner spike. The first hypothesis — that the
  notebook's fixed, global `lambda_roughness = 0.15` was over-smoothing
  these edges specifically — did NOT hold up once actually tested with
  `hdcd`'s new per-node roughness-selection option: held-out
  log-likelihood, the honest criterion, confirms 0.15 is already close to
  optimal for these edges, and a lighter penalty measurably overfits.
  The remaining mismatch looks more like a basis-expressivity limit of a
  bounded degree-4 Bernstein polynomial than a tunable smoothing bug. See
  "Known limitation" above — including the correction to the original,
  less rigorous diagnosis.
- Built the requested follow-up (a non-global, LEARNED `lambda_roughness`
  option, spec section 18) directly in the C core with its own test
  coverage, not as a notebook-only hack — see `DECISIONS.md`. It works
  correctly (verified: per-node adaptivity, no leakage into the outer
  holdout, fully backward compatible when unused) and was the right tool
  to properly check the hypothesis above, even though the answer it gave
  was "no" rather than the expected "yes."
- Followed up by raising `bernstein_degree` itself (also built directly
  in the C core, tail-dependence-gated and searched jointly with
  `lambda_roughness`, per `DECISIONS.md`): this genuinely improves
  shape-agreement with the true density on exactly the flagged edges, but
  at the cost of *worse* held-out log-likelihood at this notebook's
  sample size — a real bias-variance tradeoff between reproducing the
  true corner's shape and generalizing to new data, not a fix. See
  "Following up" above for the concrete numbers and the two remaining
  paths (more data, or a proper EVT tail-splice) logged for later.
- Growing $n$ to resolve that tradeoff was itself inconclusive (one
  realization per $n$ reversed non-monotonically) — which motivated a
  cheaper, more surgical intervention: an anisotropic roughness penalty
  that relaxes smoothing specifically near the tail-dependence corner
  instead of adding coefficients everywhere. Properly replicated (6
  seeds x 4 sample sizes), `corner_relief=0.6` improved shape-fit at
  every $n$ tested with a held-out-likelihood cost that never reached 1
  replicate standard deviation — the first of the three interventions
  tried here to clear its own pre-registered bar. See "Does corner
  relief resolve the tradeoff, properly replicated?" above.
