---
title: "Recovering a High-Dimensional Vine Copula with hdcd"
subtitle: "End-to-end validation: simulate a known mixed-family vine copula, recover it, and check the results"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 9)
```

# Purpose

This notebook is an end-to-end validation exercise for the `hdcd` library
(this repository): we

1. simulate data from a **known, high-dimensional, mixed-family vine
   copula** (Clayton, Gumbel, Frank, Gaussian, and Student-*t* pair-copulas,
   with non-uniform, heterogeneous marginals on top),
2. **recover it** with `hdcd`'s full pipeline (marginal smoothing → copula
   transform → distance-correlation dependence matrix → persistent-topology
   ordering → simulated-annealing DAG search → fixed-DAG fit),
3. time every stage, and
4. check the result three ways: does the recovered DAG match the true
   dependency structure, does its fitted score compare well to fitting the
   *true* DAG directly, and do the fitted conditional copula densities
   actually match the known, closed-form true densities?

We report timings throughout and finish with conditional density plots
comparing hdcd's nonparametric fit against ground truth.

# Setup

```{r libraries}
# This notebook lives at <repo_root>/notebooks/vine_copula_recovery.Rmd.
this_file <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NA_character_)
if (is.na(this_file) || !nzchar(this_file)) this_file <- file.path(getwd(), "vine_copula_recovery.Rmd")
repo_root <- normalizePath(file.path(dirname(this_file), ".."), mustWork = FALSE)
if (!file.exists(file.path(repo_root, "r", "DESCRIPTION"))) {
  # Fallback: assume the working directory is already the repo root.
  repo_root <- normalizePath(getwd(), mustWork = FALSE)
}
stopifnot(file.exists(file.path(repo_root, "r", "DESCRIPTION")))

if (!requireNamespace("hdcd", quietly = TRUE)) {
  message("Installing hdcd from source (", file.path(repo_root, "r"), ") ...")
  install.packages(file.path(repo_root, "r"), repos = NULL, type = "source")
}

suppressMessages({
  library(hdcd)
  library(copula)
  library(ggplot2)
  library(reshape2)
  library(knitr)
})

set.seed(20260819)
```

# A known, high-dimensional, mixed-family vine copula

## Construction

We build the ground truth as a **first-order (D-vine, tree-1) Markov-chain
copula**: variables $1, \ldots, d$ in a fixed order, where $U_{i+1}$ is
generated from a bivariate pair-copula conditional on $U_i$ alone. This is
a genuine, valid vine copula (a truncated D-vine) — the simplest member of
the D-vine family — and by mixing *different* pair-copula families and
strengths across the chain we still get a rich, high-dimensional,
non-Gaussian dependency structure.

**Why not a full multi-tree R-vine?** We tried one first. Chaining
`copula::cCopula()`'s conditional-inverse (h-function) calls through a
second tree hit a genuine numerical fragility: `uniroot()` inside the
package's Archimedean inverse-Rosenblatt implementation would occasionally
fail (`f.lower = f(lower) is NA`) once an intermediate conditional value
landed very close to the $[0,1]$ boundary after two chained inversions.
Rather than ship a demo whose *ground truth* might be silently wrong, we
kept to the tree-1 construction, which we verify below is exactly correct
(uniform marginals, matching Kendall's $\tau$), and mix five different
copula families across it for genuine complexity. Even a single-tree chain
directly exercises the part of `hdcd` we most want to validate: sparse DAG
recovery plus nonparametric conditional-copula shape estimation.

```{r vine-construction}
d <- 10L
n <- 2000L
eps <- 1e-6 # keeps chained h-function inversions off the exact [0,1] boundary

# One pair-copula per edge (i, i+1), i = 1..d-1: a deliberately varied mix.
families <- list(
  claytonCopula(3.0),   # 1 -> 2   strong lower-tail dependence
  gumbelCopula(2.2),    # 2 -> 3   upper-tail dependence
  frankCopula(5.0),     # 3 -> 4   symmetric, no tail dependence
  normalCopula(0.65),   # 4 -> 5   Gaussian
  tCopula(0.5, df = 4), # 5 -> 6   symmetric tail dependence (Student-t)
  claytonCopula(1.8),   # 6 -> 7   weaker lower-tail dependence
  gumbelCopula(3.0),    # 7 -> 8   strong upper-tail dependence
  frankCopula(3.5),     # 8 -> 9   symmetric, weaker
  normalCopula(0.45)    # 9 -> 10  Gaussian, weaker
)
family_names <- c("Clayton(3.0)", "Gumbel(2.2)", "Frank(5.0)", "Gaussian(0.65)",
                   "t(0.5, df=4)", "Clayton(1.8)", "Gumbel(3.0)", "Frank(3.5)", "Gaussian(0.45)")
stopifnot(length(families) == d - 1)

hinv <- function(u_given, w, cop) cCopula(cbind(u_given, w), copula = cop, indices = 2, inverse = TRUE)

w <- matrix(runif(n * d), ncol = d)
w <- pmin(pmax(w, eps), 1 - eps)

U <- matrix(NA_real_, n, d)
U[, 1] <- w[, 1]
for (i in 2:d) {
  u_prev <- pmin(pmax(U[, i - 1], eps), 1 - eps)
  U[, i] <- hinv(u_prev, w[, i], families[[i - 1]])
  U[, i] <- pmin(pmax(U[, i], eps), 1 - eps)
}
```

## Sanity check: is the ground truth actually correct?

A valid copula sample must have exactly Uniform(0,1) margins, and the
Kendall's $\tau$ between adjacent variables should match each pair-copula's
known $\tau$. If either check failed, nothing downstream would be worth
trusting.

```{r vine-sanity-check}
ks_pvalues <- sapply(1:d, function(j) ks.test(U[, j], "punif")$p.value)
empirical_tau <- sapply(1:(d - 1), function(i) cor(U[, i], U[, i + 1], method = "kendall"))
true_tau <- sapply(families, tau)

sanity <- data.frame(
  edge = paste0(1:(d - 1), " -> ", 2:d),
  family = family_names,
  true_kendall_tau = round(true_tau, 3),
  empirical_kendall_tau = round(empirical_tau, 3),
  marginal_KS_pvalue_ui = round(ks_pvalues[-d], 3)
)
kable(sanity, caption = "Ground-truth construction check: empirical vs. true Kendall's tau, per edge")
```

All Kolmogorov–Smirnov p-values should be comfortably large (no evidence
against uniformity), and the empirical Kendall's $\tau$ values should sit
close to their true values. This is our guarantee that the "true" copula
we're about to try to recover is exactly what we think it is.

## From copula scale to realistic, non-uniform data

`hdcd` is meant to work on **raw data**, not pre-transformed copula
coordinates — it fits its own marginals. So we push $U$ through a
heterogeneous set of marginal distributions (Normal, Exponential, and
several Normals with different means) to get a realistic-looking raw
dataset $X$, throwing away direct access to $U$ except for validation.

```{r apply-marginals}
X <- U
X[, 1] <- qnorm(U[, 1], mean = 0, sd = 2)
X[, 2] <- qexp(U[, 2], rate = 1)
for (j in 3:d) X[, j] <- qnorm(U[, j], mean = j, sd = 1)

colnames(X) <- paste0("X", 1:d)
kable(t(apply(X, 2, function(col) c(min = min(col), mean = mean(col), max = max(col)))),
      caption = "Raw data X: per-column summary (heterogeneous marginals)")
```

## What the pairwise dependence actually looks like

A quick look at three adjacent pairs, chosen to show three different
pair-copula shapes: Clayton (lower-tail clustering), Gumbel (upper-tail
clustering), and Gaussian (symmetric, elliptical).

```{r pairwise-scatter, fig.height=3.2}
pairs_df <- rbind(
  data.frame(u = U[, 1], v = U[, 2], edge = "1->2  Clayton(3.0)"),
  data.frame(u = U[, 2], v = U[, 3], edge = "2->3  Gumbel(2.2)"),
  data.frame(u = U[, 4], v = U[, 5], edge = "4->5  Gaussian(0.65)")
)
ggplot(pairs_df, aes(u, v)) +
  geom_point(alpha = 0.25, size = 0.6) +
  facet_wrap(~edge) +
  coord_equal() +
  theme_minimal() +
  labs(title = "Copula-scale pairwise dependence for three representative edges",
       x = expression(U[i]), y = expression(U[i+1]))
```

# Fitting hdcd

We now forget we ever saw `U` or the true DAG, and hand `hdcd` only the raw
data `X`. We run the pipeline stage by stage (rather than calling the
single `hdcd_fit()` wrapper) purely so we can time each stage separately;
`hdcd_fit()` performs exactly the same steps internally.

```{r fit-stages}
timings <- list()
tic <- function() Sys.time()
toc <- function(t0, label) {
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
  timings[[label]] <<- elapsed
  invisible(elapsed)
}

# ---- Stage 1: marginal fitting (spec section 2) ----
t0 <- tic()
marginals <- lapply(1:d, function(j) hdcd:::.new_marginal(X[, j]))
toc(t0, "1. Marginal fitting (d columns)")

# ---- Stage 2: copula transform (spec section 4) ----
t0 <- tic()
U_hat <- matrix(NA_real_, n, d)
for (j in 1:d) U_hat[, j] <- hdcd:::.transform_to_copula(marginals[[j]], X[, j])
toc(t0, "2. Copula transform (d columns)")

# ---- Stage 3: dependence matrix (spec section 5) ----
t0 <- tic()
dm_ext <- hdcd:::.compute_dependence_matrix(U_hat, n, d)
toc(t0, "3. Distance-correlation dependence matrix")

# ---- Stage 4: persistent-topology ordering (spec section 6) ----
t0 <- tic()
topo_ext <- hdcd:::.compute_topology(dm_ext)
ordering <- hdcd:::.topology_ordering(topo_ext)
toc(t0, "4. Persistent-topology ordering")

# ---- Stage 5: simulated-annealing DAG search (spec sections 7, 17) ----
max_parents <- 2L
lambda_edge <- 0.05
bernstein_degree <- 4L
lambda_roughness <- 0.15
holdout_fraction <- 0.25
seed <- 1L
local_seed <- seed + 1L

t0 <- tic()
annealed <- hdcd:::.run_annealing_c(
  U_hat, n, d, ordering, max_parents, lambda_edge,
  bernstein_degree, lambda_roughness, holdout_fraction, local_seed,
  initial_temperature = 0.5, cooling_rate = 0.95,
  max_iterations = 600L, restarts = 3L,
  p_add = 1.0, p_remove = 1.0, p_swap = 1.0, anneal_seed = seed
)
toc(t0, "5. Simulated-annealing DAG search (1800 total proposals)")

# ---- Stage 6: final fixed-DAG fit (spec sections 14, 28) ----
t0 <- tic()
dag_fit_ext <- hdcd:::.dag_fit_c(U_hat, n, d, annealed$dag, bernstein_degree,
                                  lambda_roughness, holdout_fraction, local_seed)
toc(t0, "6. Final fixed-DAG fit")

timing_df <- data.frame(
  stage = names(timings),
  seconds = round(unlist(timings), 3)
)
timing_df <- rbind(timing_df, data.frame(stage = "TOTAL", seconds = round(sum(unlist(timings)), 3)))
kable(timing_df, caption = sprintf("Stage-by-stage timing (d=%d, n=%d)", d, n))
```

```{r assemble-model}
# Assemble the same hdcd_model structure hdcd_fit() would return, so the
# rest of the notebook can use the normal public API.
model <- list(
  marginals = marginals, dag = annealed$dag, dag_fit = dag_fit_ext,
  dependence_matrix_ext = dm_ext, topology_ext = topo_ext, d = d,
  max_parents = max_parents, bernstein_degree = bernstein_degree,
  lambda_roughness = lambda_roughness, holdout_fraction = holdout_fraction,
  local_seed = local_seed, best_score = annealed$score,
  score_trace = annealed$score_trace, accepted_trace = annealed$accepted_trace,
  acceptance_rate = annealed$acceptance_rate, U = U_hat, X = X
)
class(model) <- "hdcd_model"
print(model)
```

# Did it recover the structure?

## Recovered vs. true DAG

```{r dag-comparison}
recovered_edges <- hdcd_dag(model)
true_edges <- cbind(1:(d - 1), 2:d)

edge_key <- function(e) paste(pmin(e[, 1], e[, 2]), pmax(e[, 1], e[, 2]), sep = "-")
recovered_skeleton <- edge_key(recovered_edges)
true_skeleton <- edge_key(true_edges)

comparison <- data.frame(
  true_edge = paste0(true_edges[, 1], " -> ", true_edges[, 2]),
  recovered_as = sapply(seq_len(nrow(true_edges)), function(i) {
    key <- true_skeleton[i]
    idx <- which(recovered_skeleton == key)
    if (length(idx) == 0) return("MISSED")
    re <- recovered_edges[idx, ]
    if (re[1] == true_edges[i, 1]) "exact match" else sprintf("reversed (%d -> %d)", re[1], re[2])
  })
)
kable(comparison, caption = "Every true edge: was it recovered, and in the correct direction?")

n_exact <- sum(comparison$recovered_as == "exact match")
n_reversed <- sum(grepl("reversed", comparison$recovered_as))
n_missed <- sum(comparison$recovered_as == "MISSED")
n_spurious <- nrow(recovered_edges) - (n_exact + n_reversed)

cat(sprintf(
  "Skeleton edges recovered: %d / %d (%d exact direction, %d reversed). Spurious extra edges: %d.\n",
  n_exact + n_reversed, nrow(true_edges), n_exact, n_reversed, n_spurious
))
```

A reversed edge is not a bug: `hdcd`'s reference DAG is an optimized
**statistical factorization for density estimation**, not a causal claim
(this is stated explicitly throughout the library's own documentation and
tested directly in Milestone 9). For a chain, the pairwise dependence
between $U_i$ and $U_{i+1}$ is symmetric regardless of which one is
labelled "parent," so a direction flip on one edge — while everything
else matches — is exactly the kind of statistically-near-equivalent
alternative factorization the annealing search is allowed to land on.

## Dependence matrix and persistent-topology ordering

```{r dependence-heatmap, fig.height=5}
dm <- hdcd_dependence_matrix(model)
dm_long <- melt(dm, varnames = c("row", "col"), value.name = "dCor")
ggplot(dm_long, aes(col, row, fill = dCor)) +
  geom_tile() +
  scale_fill_viridis_c(limits = c(0, 1)) +
  scale_y_reverse() +
  coord_equal() +
  theme_minimal() +
  labs(title = "Fitted pairwise distance-correlation matrix", x = "column", y = "column")
```

The chain structure is clearly visible as a bright band along the
super-/sub-diagonal (adjacent variables are strongly dependent), fading
away from it — exactly the pattern a first-order Markov chain should
produce.

```{r ordering}
cat("Persistent-topology ordering:", hdcd_ordering(model), "\n")
cat("(true generative order was 1, 2, ..., 10 -- any ordering consistent\n",
    " with the chain's actual dependency strengths is a valid recovery.)\n")
```

## Did the annealing search actually converge?

```{r annealing-trace, fig.height=3.5}
trace_df <- data.frame(iteration = seq_along(model$score_trace), score = model$score_trace)
ggplot(trace_df, aes(iteration, score)) +
  geom_line(alpha = 0.5) +
  geom_hline(yintercept = model$best_score, color = "firebrick", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Simulated-annealing score trace (J(G) after each iteration)",
       subtitle = sprintf("acceptance rate = %.2f; dashed line = best score found (%.4f)",
                           model$acceptance_rate, model$best_score),
       x = "iteration (across all restarts)", y = "J(G)  [lower is better]")
```

$J(G)$ should trend down and then flatten out as the search converges on a
good structure — a rising-then-flat noisy trace, not one that's still
improving steeply at the end, is what confirms the iteration budget was
sufficient.

# How good is the recovered DAG's fit, really?

Structural mismatches on individual edges don't necessarily mean a worse
*fit*. We can check directly: fit the true DAG explicitly with
`hdcd_fit_dag()` (over the exact same training data) and compare held-out
KL scores with `hdcd_score_dag()` (spec section 19).

```{r true-dag-fit}
t0 <- tic()
true_fit <- hdcd_fit_dag(model, true_edges)
true_fit_time <- toc(t0, "True-DAG fit_dag() (for comparison only)")

delta_kl <- hdcd_score_dag(model, true_fit)
cat(sprintf("fit_dag(true DAG) took %.2f seconds\n", true_fit_time))
cat(sprintf(
  "Delta_KL(true DAG vs. recovered/annealed DAG) = %.4f\n(negative means the TRUE dag scores BETTER; near 0 means the annealed DAG is nearly as good)\n",
  delta_kl
))
```

# Do the fitted conditional densities actually match?

This is the most direct test of `hdcd`'s core nonparametric machinery
(the centered Bernstein tensor kernel plus Sinkhorn normalization,
Milestones 5-6): for every true edge, does `hdcd`'s fitted conditional
copula density $\hat c(u \mid z)$ — evaluated from the **true-DAG fit**,
so this is purely about density-shape accuracy, independent of the
discrete structure-search question above — match the known, closed-form
conditional density of the true pair-copula?

```{r conditional-density-grid, fig.height=14}
u_grid <- seq(0.02, 0.98, length.out = 60)
q_values <- c(0.1, 0.5, 0.9)

cond_rows <- list()
for (i in seq_len(d - 1)) {
  node <- i + 1L # child column (1-indexed), matches true_fit's node numbering
  for (q in q_values) {
    fitted <- exp(hdcd:::.local_fit_conditional_log_density(true_fit$dag_fit, node, u_grid, q))
    true_dens <- dCopula(cbind(rep(q, length(u_grid)), u_grid), families[[i]])
    cond_rows[[length(cond_rows) + 1]] <- data.frame(
      edge = sprintf("%d: %d->%d  %s", i, i, i + 1, family_names[i]),
      q = factor(sprintf("z = %.1f", q), levels = sprintf("z = %.1f", q_values)),
      u = u_grid,
      density = c(fitted, true_dens),
      source = rep(c("hdcd (fitted)", "true"), each = length(u_grid))
    )
  }
}
cond_df <- do.call(rbind, cond_rows)

ggplot(cond_df, aes(u, density, color = source, linetype = source)) +
  geom_line(linewidth = 0.7) +
  facet_grid(edge ~ q, scales = "free_y") +
  scale_color_manual(values = c("hdcd (fitted)" = "steelblue", "true" = "black")) +
  theme_minimal(base_size = 9) +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), legend.position = "top") +
  labs(title = "Fitted vs. true conditional copula density c(u | z) for every true edge",
       x = expression(u[child]), y = "density")
```

## Summary agreement metrics

```{r conditional-density-summary}
agreement <- do.call(rbind, lapply(seq_len(d - 1), function(i) {
  node <- i + 1L
  per_q <- sapply(q_values, function(q) {
    fitted <- exp(hdcd:::.local_fit_conditional_log_density(true_fit$dag_fit, node, u_grid, q))
    true_dens <- dCopula(cbind(rep(q, length(u_grid)), u_grid), families[[i]])
    c(cor = cor(fitted, true_dens), rmse = sqrt(mean((fitted - true_dens)^2)))
  })
  data.frame(
    edge = sprintf("%d->%d", i, i + 1), family = family_names[i],
    mean_correlation = round(mean(per_q["cor", ]), 3),
    mean_rmse = round(mean(per_q["rmse", ]), 3)
  )
}))
kable(agreement, caption = "Per-edge agreement between hdcd's fitted density and the true density, averaged over z in {0.1, 0.5, 0.9}")

cat(sprintf("\nOverall mean correlation across all edges/conditioning values: %.3f\n",
            mean(agreement$mean_correlation)))
```

The per-edge table above is not uniform: the two smooth, no-corner-singularity
families (Frank, Gaussian) sit at correlation $\geq 0.97$, while the three
edges built from **tail-dependent** families (Clayton at edges 1 and 6,
Gumbel at edges 2 and 7) sit noticeably lower (0.90-0.96). That is not
random scatter — it is a real, systematic shape mismatch, worth
understanding rather than averaging away into the single headline number
above.

## Known limitation: sharp tail-dependence corners are under-fit

Look closely at the row-1 and row-6 panels (Clayton) at $z=0.1$, and the
row-7 panel (Gumbel) at $z=0.9$, in the density grid above: the **true**
density has a sharp, narrow spike right at the copula corner (Clayton's
lower-tail dependence concentrates mass as $u, z \to 0$; Gumbel's
upper-tail dependence does the same as $u, z \to 1$), while `hdcd`'s
**fitted** curve is visibly flatter and shorter there, even though it
tracks the true curve well everywhere else. This is the actual thing that
prompted a closer look at "densities that don't line up."

The first hypothesis was that a single, hand-picked, *global*
`lambda_roughness = 0.15` — applied identically to all nine edges — was
over-smoothing specifically these sharply-peaked edges, and that a
lighter, per-edge penalty would fit them better. `hdcd` now supports
exactly that as an option (`lambda_roughness_grid`, see `DECISIONS.md`):
a small validation grid, searched independently per node via an *inner*
train/validation split that never touches the outer holdout, entirely
outside the annealing search (spec section 18). Applying it directly
tests the hypothesis rather than just asserting it.

```{r roughness-grid-fit}
roughness_grid <- c(0.02, 0.05, 0.1, 0.15, 0.3, 0.6, 1.0)
t0 <- tic()
auto_fit <- hdcd:::.dag_fit_c(model$U, n, d, true_fit$dag, bernstein_degree, lambda_roughness,
                               holdout_fraction, local_seed,
                               lambda_roughness_grid = roughness_grid,
                               roughness_validation_fraction = 0.3)
auto_fit_time <- toc(t0, "Per-node auto lambda_roughness fit_dag() (diagnostic only)")

selected <- do.call(rbind, lapply(seq_len(d - 1), function(i) {
  node <- i + 1L
  data.frame(
    edge = sprintf("%d->%d", i, i + 1), family = family_names[i],
    selected_lambda = hdcd:::.local_fit_selected_lambda_roughness(auto_fit, node),
    holdout_score_auto = round(hdcd:::.dag_fit_holdout_scores(auto_fit, d)[node], 4),
    holdout_score_fixed_0.15 = round(hdcd:::.dag_fit_holdout_scores(true_fit$dag_fit, d)[node], 4)
  )
}))
kable(selected, caption = sprintf(
  "Per-node lambda_roughness selected from {%s} by inner-validation, vs. this notebook's fixed 0.15",
  paste(roughness_grid, collapse = ", ")
))
cat(sprintf("Delta_KL(fixed@0.15 vs. auto-selected) = %.4f\n",
            hdcd:::.dag_fit_kl_difference(true_fit$dag_fit, auto_fit)))
```

**The hypothesis does not survive contact with the data.** Every edge —
Clayton and Gumbel included — selects a lambda at or above this
notebook's original 0.15, not below it; several land higher (0.3). A
finer, continuous sweep (fixed `lambda_roughness`, no grid search)
confirms this is a genuine interior optimum, not an artifact of the
grid's spacing:

```{r roughness-fine-sweep, fig.height=3.2}
fine_lambdas <- c(0.01, 0.02, 0.05, 0.1, 0.15, 0.3, 0.6, 1.0, 2.0)
fine_sweep <- do.call(rbind, lapply(c(edge1 = 2L, edge7 = 8L, edge3 = 4L), function(node) {
  do.call(rbind, lapply(fine_lambdas, function(lr) {
    fit <- hdcd:::.dag_fit_c(model$U, n, d, true_fit$dag, bernstein_degree, lr, holdout_fraction, local_seed)
    data.frame(edge = c(`2` = "1->2 Clayton(3.0)", `8` = "7->8 Gumbel(3.0)", `4` = "3->4 Frank(5.0)")[[as.character(node)]],
               lambda_roughness = lr,
               holdout_score = hdcd:::.dag_fit_holdout_scores(fit, d)[node])
  }))
}))
ggplot(fine_sweep, aes(lambda_roughness, holdout_score)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = 0.15, linetype = "dashed", color = "firebrick") +
  scale_x_log10() +
  facet_wrap(~edge, scales = "free_y") +
  theme_minimal() +
  labs(title = "Held-out score vs. lambda_roughness (log scale), fixed bernstein_degree",
       subtitle = "dashed line = this notebook's fixed 0.15",
       x = "lambda_roughness (log scale)", y = "held-out log-density score")
```

The same interior-optimum shape shows up for the Clayton edge, the
Gumbel edge, and even the smooth Frank edge (which has no corner at
all) — none of them want much lighter regularization than 0.15, and all
of them clearly punish going much lighter. `Delta_KL` in the table above
is slightly *positive* for the auto-selected fit relative to the fixed
default: per-node selection pays a real cost here (each candidate is fit
on a smaller *inner*-train subset, which is why it exists to begin with
— see `DECISIONS.md`), and on this dataset that cost isn't repaid,
because 0.15 was already close to optimal everywhere it was tested.

**Revised conclusion.** The original diagnosis — eyeballing the fitted
curve against the true curve and concluding "a lighter penalty would fit
this corner better" — was measuring agreement with a density that is only
knowable here because this is a synthetic exercise with ground truth.
*Held-out log-likelihood*, the criterion `hdcd` (and any real, non-
synthetic application) actually optimizes, disagrees: it says 0.15 is
already a good choice for these edges, lighter regularization measurably
overfits, and the visual corner-flattening is not primarily a smoothing
problem. What's left is more consistent with a basis-expressivity
limit — a degree-4 centered Bernstein tensor is a smooth, globally
bounded polynomial surface, and no amount of retuning `lambda_roughness`
lets a bounded polynomial reproduce an unbounded corner singularity. This
is a more honest, if less tidy, takeaway than the first pass reached, and
exactly what building the feature properly (rather than assuming the fix)
was for.

## Following up: does raising bernstein_degree, done properly, help?

The natural next lever, and a genuinely different question from the
`lambda_roughness` one above: is a degree-4 centered Bernstein tensor
just too coarse a *basis* to represent a sharp corner, independent of how
it's regularized? `hdcd` now supports raising `bernstein_degree`
per-node too (`bernstein_degree_grid`), gated by an empirical
tail-dependence coefficient (a standard nonparametric EVT/copula
estimator — see `DECISIONS.md`'s "tail-dependence-informed
`bernstein_degree` selection") so the extra flexibility (and its cost) is
only spent on edges that actually show tail dependence, and searched
*jointly* with `lambda_roughness` this time — the earlier, very first
degree check in this notebook held lambda fixed while raising degree,
which confounded "does more basis flexibility help" with "is it
regularized correctly for its new size."

```{r degree-grid-fit}
degree_grid <- c(4L, 6L, 8L, 10L)
lambda_grid_2 <- c(0.05, 0.1, 0.15, 0.3)
t0 <- tic()
degree_fit <- hdcd:::.dag_fit_c(model$U, n, d, true_fit$dag, bernstein_degree, lambda_roughness,
                                 holdout_fraction, local_seed,
                                 lambda_roughness_grid = lambda_grid_2, roughness_validation_fraction = 0.3,
                                 bernstein_degree_grid = degree_grid, tail_dependence_gate = 0.05)
degree_fit_time <- toc(t0, "Tail-dependence-gated joint (degree, lambda) fit_dag() (diagnostic only)")

degree_summary <- do.call(rbind, lapply(seq_len(d - 1), function(i) {
  node <- i + 1L
  q <- if (i %in% c(2, 7)) 0.9 else 0.1 # Gumbel edges: upper tail; else lower
  fitted_fixed <- exp(hdcd:::.local_fit_conditional_log_density(true_fit$dag_fit, node, u_grid, q))
  fitted_auto <- exp(hdcd:::.local_fit_conditional_log_density(degree_fit, node, u_grid, q))
  true_dens <- dCopula(cbind(rep(q, length(u_grid)), u_grid), families[[i]])
  data.frame(
    edge = sprintf("%d->%d", i, i + 1), family = family_names[i],
    tail_dependence = round(hdcd:::.local_fit_max_tail_dependence(degree_fit, node), 3),
    selected_degree = hdcd:::.local_fit_selected_bernstein_degree(degree_fit, node),
    selected_lambda = hdcd:::.local_fit_selected_lambda_roughness(degree_fit, node),
    cor_degree4_fixed = round(cor(fitted_fixed, true_dens), 3),
    cor_degree_auto = round(cor(fitted_auto, true_dens), 3)
  )
}))
kable(degree_summary, caption = sprintf(
  "Tail-dependence-gated joint (degree in {%s}, lambda in {%s}) selection vs. the fixed degree=4/lambda=0.15 baseline",
  paste(degree_grid, collapse = ", "), paste(lambda_grid_2, collapse = ", ")
))
cat(sprintf("Delta_KL(fixed baseline vs. degree-auto fit) = %.4f\n",
            hdcd:::.dag_fit_kl_difference(true_fit$dag_fit, degree_fit)))
```

Two things happen at once, and they point in opposite directions:

- **The tail-dependence diagnostic itself is doing its job.** The
  measured coefficient is high for exactly the edges flagged visually
  (Clayton, Gumbel) and low for the smooth ones (Frank, Gaussian) — this
  is the same corner-concentration property being picked up
  independently, by an estimator that has never seen a fitted density
  curve.
- **Shape-correlation to the true density DOES improve on the flagged
  edges** once degree is raised jointly with a re-tuned (lighter)
  lambda — unlike the very first, unfair degree-only check earlier in
  this notebook. Both Clayton edges and both Gumbel edges gain several
  points of correlation.
- **But `Delta_KL` goes the wrong way**: the auto (higher-degree)
  fit's *held-out log-likelihood* is measurably worse than the plain
  degree=4/lambda=0.15 baseline, not better. A degree-10 tensor has
  $11 \times 11 = 121$ coefficients per edge instead of $5\times5=25$ at
  degree 4 — roughly 5x the parameters, fit from a training set this
  notebook never grew to match. It reproduces the true curve's *shape*
  more closely while generalizing *worse* to new data at $n=2000$: the
  same shape-vs-likelihood tension the `lambda_roughness` investigation
  surfaced, now showing up again one level up, in the choice of basis
  size.

**Where this leaves the original question.** Raising `bernstein_degree`
is not a free win either, at least not at this sample size: it can
literally trade held-out predictive accuracy for a prettier picture of a
density nobody could observe in a real (non-synthetic) application. Two
honest paths forward, both logged in `DECISIONS.md` rather than picked
here: grow $n$ until the extra flexibility pays for itself (this
notebook fixes $n=2000$ throughout — a genuinely separate experiment),
or reach for the heavier, more targeted fix — a true EVT tail-splice
that adds a handful of *parametric* tail-behavior parameters exactly
where the corner is, instead of dozens more general-purpose polynomial
coefficients everywhere.

## Does more data resolve the bias-variance tradeoff?

Picking up the first of those two paths directly: does growing $n$ make
the higher-`bernstein_degree` fit actually pay off in held-out
log-likelihood, not just in shape-correlation? We reran the exact same
tail-dependence-gated joint (degree, lambda) search — degree in
$\{4,6,8,10\}$, lambda in $\{0.05,0.1,0.15,0.3\}$, gate $0.05$ — against
the same true DAG, at $n \in \{2000, 4000, 8000, 16000\}$, tracking the
two most severely under-fit edges (1->2 Clayton, 7->8 Gumbel) plus the
overall (all 9 edges) `Delta_KL`.

This sweep is genuinely expensive — the joint search's cost grows worse
than linearly in $n$ (19s at $n=2000$ vs. 879s at $n=16000$ for the
9-edge search alone, on top of copula-scale data generation itself
taking up to 84s at the largest $n$), roughly half an hour in total — so
it was run once, standalone
(`notebooks/n_sweep_experiment.R`, saved to `notebooks/n_sweep_results.csv`),
rather than re-executed on every render of this notebook.

```{r n-sweep-load}
sweep_results <- read.csv("n_sweep_results.csv")
kable(sweep_results[, c("n", "edge", "family", "selected_degree", "selected_lambda",
                         "cor_fixed", "cor_auto", "delta_kl_total")],
      digits = 3, row.names = FALSE,
      caption = "Tail-dependence-gated joint search vs. sample size (one realization per n; see notebooks/n_sweep_experiment.R)")
```

```{r n-sweep-plot, fig.height=3.2}
kl_by_n <- unique(sweep_results[, c("n", "delta_kl_total")])
p1 <- ggplot(kl_by_n, aes(n, delta_kl_total)) +
  geom_line() + geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Delta_KL(fixed vs. auto), all 9 edges", subtitle = "0 = tied; positive = fixed baseline still better",
       x = "n (log scale)", y = "Delta_KL")

cor_long <- reshape2::melt(sweep_results[, c("n", "edge", "cor_fixed", "cor_auto")],
                            id.vars = c("n", "edge"), variable.name = "fit", value.name = "correlation")
p2 <- ggplot(cor_long, aes(n, correlation, color = fit, linetype = edge)) +
  geom_line() + geom_point(size = 1.5) +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Shape-correlation to true density", x = "n (log scale)", y = "correlation")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

**The first three points looked like a clean story, and then it wasn't.**
`Delta_KL` fell from 0.252 (n=2000) to 0.203 (n=4000) to 0.099 (n=8000) —
each doubling roughly halving the gap, exactly what "the tradeoff is a
data-size effect" would predict, seemingly on track to cross zero
somewhere past $n=16000$. It didn't: at $n=16000$, `Delta_KL` jumped back
up to 0.319, worse than every smaller $n$ tested, including $n=2000$.

Meanwhile the *shape*-correlation numbers stayed almost boringly stable
across the whole sweep — `cor_auto` for 1->2 sits in a narrow 0.80-0.82
band and `cor_auto` for 7->8 in 0.83-0.86, at every single $n$, always a
few points above the matching `cor_fixed`. And the search picked the
grid's most flexible option (degree 10, lambda 0.05 — literally the
edge of the search space) every single time, at every $n$, never
settling partway into the grid as more data arrived.

**Why this is a real finding, not a wasted run.** Three things are true
at once here, and they don't resolve into a tidy story:

1. The search's own selection criterion (inner-validation score) always
   wants maximum flexibility, at every sample size tested — the grid's
   boundary never stopped being attractive, which on its own is a mild
   red flag that the grid itself may be too narrow to see where the
   optimum actually sits.
2. Shape-correlation to the true density is a stable, small, consistent
   win for the higher-degree fit across the whole range tested.
3. Held-out log-likelihood — the metric that actually matters — is
   noisy enough, and moved by enough else besides raw sample size, that
   four points spanning an 8x range in $n$ produced a non-monotonic
   trend with a reversal at the *largest* $n$, not the smallest.

Point 3 matters most. **This sweep is a single realization per $n$, not
a replicated design**: each $n$ draws a fresh, independent copula sample
rather than growing one fixed dataset, so sample-to-sample variability
in exactly which points land in which train/inner-validation/holdout
split is completely confounded with the effect of $n$ itself. A
1-point-per-$n$ curve cannot distinguish "the tradeoff doesn't resolve
monotonically" from "we got an unlucky draw at $n=16000$." Properly
separating those would need several replicate seeds at every $n$ and a
confidence band — and at roughly 15 minutes of compute for a *single*
$n=16000$ replicate, getting even 5 replicates there costs over an hour,
for one edge of one synthetic experiment.

That gap — between "the question is well-posed" and "answering it
rigorously is affordable" — is itself worth sitting with. Confirming
whether more data resolves a flexibility/generalization tradeoff in a
high-dimensional, heavy-tailed dependence model is not just a modeling
question; empirically *checking* the answer runs into the same
high-dimensional, heavy-tailed cost structure the modeling problem
started with. Logged in `DECISIONS.md` as an open question rather than
resolved here: the honest result of this sweep is "inconclusive, and
expensive to make conclusive," not "yes" or "no."

## Does corner relief resolve the tradeoff, properly replicated?

The `bernstein_degree_grid` route added *more coefficients everywhere*
and paid a generalization cost for it. A cheaper, more surgical
alternative discussed as a follow-up: relax the roughness penalty
specifically near the Bernstein coefficient grid's four CORNERS — where
$u$ and $z$ are both extreme simultaneously, exactly where tail
dependence concentrates — instead of raising the coefficient count.
`hdcd` now supports this directly
(`corner_relief`, see `DECISIONS.md`'s "anisotropic (corner-relaxed)
roughness penalty"): one scalar in $[0,1)$, applied identically inside
every Theta fit, at zero extra cost per fit (unlike a grid search).

That last property — no extra fit calls — makes something the earlier
`bernstein_degree_grid` sweep couldn't afford actually affordable here:
**real replication.** Six independent seeds at every $n \in \{2000,
4000, 8000, 16000\}$, for `corner_relief` $\in \{0, 0.3, 0.6, 0.8\}$
(fixed `bernstein_degree=4`, `lambda_roughness=0.15` throughout, so
`corner_relief` is the only thing varying), on the same two worst edges
(1->2 Clayton, 7->8 Gumbel). Standalone script
`notebooks/corner_relief_sweep_experiment.R` (~48 minutes total),
results saved to `notebooks/corner_relief_sweep_results.csv` and loaded
here, not recomputed.

```{r corner-relief-sweep-load}
cr_results <- read.csv("corner_relief_sweep_results.csv")
cr_agg <- aggregate(
  cbind(cor_true, delta_kl_total) ~ n + corner_relief, data = cr_results,
  FUN = function(x) c(mean = mean(x), sd = sd(x))
)
cr_agg <- do.call(data.frame, cr_agg)
names(cr_agg) <- c("n", "corner_relief", "cor_mean", "cor_sd", "delta_kl_mean", "delta_kl_sd")
cr_agg$z_score <- ifelse(cr_agg$delta_kl_sd > 0, cr_agg$delta_kl_mean / cr_agg$delta_kl_sd, NA)
kable(cr_agg, digits = 3, row.names = FALSE,
      caption = "Mean +/- SD over 6 replicates x 2 edges (12 observations per row). z_score = Delta_KL mean / Delta_KL SD.")
```

```{r corner-relief-sweep-plot, fig.height=3.2}
# Error bars plot +/- 1 SD, matching the z-score criterion above
# (delta_kl_mean / delta_kl_sd) EXACTLY -- not +/- SE (sd/sqrt(12)),
# which would make the bars ~3.5x narrower than what the pre-registered
# "should not sit outside 1-2 SD" bar in DECISIONS.md actually judges
# the result against, understating the real uncertainty in the plot
# relative to the text.
p1 <- ggplot(cr_agg, aes(n, cor_mean, color = factor(corner_relief))) +
  geom_line() +
  geom_pointrange(aes(ymin = cor_mean - cor_sd, ymax = cor_mean + cor_sd)) +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Shape-correlation to true density", subtitle = "mean +/- 1 SD, 12 obs/cell",
       x = "n (log scale)", y = "correlation", color = "corner_relief")

p2 <- ggplot(subset(cr_agg, corner_relief > 0), aes(n, delta_kl_mean, color = factor(corner_relief))) +
  geom_line() +
  geom_pointrange(aes(ymin = delta_kl_mean - delta_kl_sd, ymax = delta_kl_mean + delta_kl_sd)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  scale_x_log10() +
  theme_minimal() +
  labs(title = "Delta_KL vs. corner_relief=0 baseline", subtitle = "mean +/- 1 SD; 0 = tied with baseline",
       x = "n (log scale)", y = "Delta_KL", color = "corner_relief")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

**Look at the error bars before reading anything into the lines.** These
are $\pm 1$ SD, matching the z-score criterion exactly (not $\pm$ SE,
which would make the bars $\sqrt{12} \approx 3.5\times$ narrower and
visually overstate how separated these lines really are). At that honest
scale, every `corner_relief` value's error bar overlaps every other
value's, and every `Delta_KL` error bar overlaps zero, at every single
$n$. Taken one point at a time, **no individual (n, corner_relief) cell
is distinguishable from any other** — this is not a plot where a line
"clearly wins."

What the plot does show, and what the pre-registered criterion was
actually designed to detect, is a *pattern in the means* rather than any
single significant point: `corner_relief=0.6`'s mean shape-correlation is
the highest (or tied-highest) of the four values at all four independent
$n$ draws, and its mean `Delta_KL` cost stays small enough, relative to
its own spread, to never exceed 1 SD (z-scores 0.14–0.98) — a weaker,
more honest form of evidence than "the point estimate beats baseline
outside its error bar," but a real one: with $12$ independent
observations per cell and substantial noise, four-out-of-four consistent
directional wins is not what pure chance typically produces, even though
no *individual* comparison clears significance on its own. Contrast this
with the `bernstein_degree_grid` sweep two sections up, which had no
replication at all — not "the effect is small relative to noise we
measured," but literally no way to tell signal from noise, at any $n$.

Two things not to paper over. First, the z-score drifts upward with $n$
(0.39, 0.14, 0.86, 0.98) — still under 1 SD throughout, but trending
toward it, not away from it; six replicates rules out "the degree-grid's
$n=16000$ reversal was real," but does not rule out a genuine,
slowly-growing `corner_relief` cost past the range tested here. Second,
and more important given how wide these bars are: this result is
*directional, weak evidence*, not a demonstrated effect. It would take
materially more replicates (or a purpose-built sequential design) before
any single `Delta_KL` estimate here could stand on its own without
leaning on the four-way consistency argument.

**Bottom line:** of the three interventions tried in this notebook for
the Clayton/Gumbel corner under-fit — a validated `lambda_roughness`
(ruled out: the fixed default was already near-optimal), a validated
`bernstein_degree` (real shape gain, but a real and non-monotonic
likelihood cost that didn't resolve with more data), and now
`corner_relief` — this is the first to clear its own pre-registered bar,
but "cleared a lenient, pre-registered bar via a consistent-direction
argument across four noisy points" is the accurate description, not
"proven to work." `corner_relief=0.6` is not adopted as `hdcd`'s new
default here (that is a decision for whoever maintains this library
going forward, not made unilaterally in a validation notebook) — it is
a lead worth a properly powered follow-up, not a settled result.


## Does a true EVT tail-splice, with the objectively correct family, finally fix the corner?

Every intervention tried so far — `lambda_roughness_grid`,
`bernstein_degree_grid`, `corner_relief` — worked entirely within the
nonparametric Bernstein/Sinkhorn machinery: more or fewer polynomial
coefficients, regularized more or less, everywhere or just at the
corner. None of them ever introduced an actual *parametric copula
family*, even though the ground truth here literally is one (Clayton,
Gumbel, ...) on the flagged edges. `hdcd` briefly supported this
directly: a copula-level EVT tail-splice (`evt_splice_gate`/
`evt_splice_bandwidth`) that, on edges whose empirical tail-dependence
coefficient cleared a gate, fit a genuine Clayton or Gumbel density by
MLE and blended it into the raw Bernstein kernel near the corresponding
corner — before Sinkhorn normalized the result, so no hand-built
continuity matching was needed; Sinkhorn already turns any positive raw
kernel into a valid conditional copula density.

**This feature has since been removed from `hdcd`.** Assuming a
parametric copula family for the dependence structure is outside this
library's design: spec section 3's "Extreme-Value Tail Extension" is a
*marginal*-level provision (splicing generalized Pareto tails onto each
dimension's univariate CDF), and says nothing about the joint copula
density, which the centered-Bernstein-tensor-plus-Sinkhorn machinery
(spec section 9 onward) is deliberately designed to estimate *without*
assuming any family. This section is kept, unlike the removed API,
because the finding is still a real and informative part of this
investigation — it reports results from a saved run of
`notebooks/evt_splice_experiment.R`, executed once while the feature
still existed in the C core, rather than calling the (now-removed)
live API. See `DECISIONS.md`'s "copula-level EVT tail-splice" entry for
the full history, including the removal decision.

```{r evt-splice-load}
evt_results <- readRDS("evt_splice_experiment_results.rds")
```

The first attempt used a permissive gate (`evt_splice_gate = 0.1`) and a
bandwidth borrowed, unjustified, from `corner_relief` (`0.15`):

```{r evt-splice-miscalibrated}
kable(evt_results$evt_wide_summary,
      caption = "evt_splice_gate=0.1, evt_splice_bandwidth=0.15 (miscalibrated first attempt)")
```

Two problems, visible directly in that table. First, the gate is far too
permissive: it splices Clayton or Gumbel onto every edge, including
Frank, Gaussian, and t edges with no genuine tail dependence — and it
actively hurts those (the Frank(3.5) edge's correlation drops from 0.999
to well below that). Second, even on the edges where the splice picks
the objectively *correct* family — Clayton on both true-Clayton edges,
Gumbel on both true-Gumbel edges — shape-correlation to the true density
improves on only one of those four edges and gets *worse* on the other
three, despite the MLE-fit theta being accurate (e.g. 2.93 vs. a true
3.0). Using the right family with an accurate parameter did not, on its
own, help.

Root cause: the blend is a weighted *geometric* mean of the raw
(unnormalized) Bernstein kernel with the already-normalized parametric
copula density. At `bandwidth=0.15` the parametric term dominates well
outside the actual corner, into the bulk of the square where the
Bernstein tensor already fits well by itself — the resulting scale
mismatch between an unnormalized and a normalized term degrades the
Sinkhorn-normalized fit even where family selection was correct. Raising
the gate to `0.5` (matching `bernstein_degree_grid`'s own gate) and
tightening the bandwidth to confine the splice to the actual corner
fixes both problems at once:

```{r evt-splice-calibrated}
kable(evt_results$evt_summary, caption = "evt_splice_gate=0.5, evt_splice_bandwidth=0.08 (calibrated)")
cat(sprintf("Delta_KL(fixed baseline vs. evt-splice fit) = %.4f\n", evt_results$delta_kl_evt))
```

The correlation numbers above are a summary statistic; the actual shape
of the corner is the thing that matters. Plotting the three curves
directly, at the same corner slice (`q`) used in the table, for the
four genuinely tail-dependent edges:

```{r evt-splice-density-plot, fig.height=8}
evt_cond_df <- evt_results$evt_cond_df

ggplot(evt_cond_df, aes(u, density, color = source, linetype = source)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~edge, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c("true" = "black", "hdcd (unspliced)" = "steelblue",
                                 "hdcd (evt-spliced)" = "firebrick")) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "top") +
  labs(title = "True vs. unspliced vs. EVT-spliced conditional density at the tail-dependence corner",
       x = expression(u[child]), y = "density")
```

**On this slice the two curves really are nearly indistinguishable —
but that is a property of *where this slice sits*, not of whether the
splice does anything at all.** `q = 0.1`/`0.9` is the slice used
throughout this notebook (including the "Known limitation" plot above),
but the splice's corner weight
$w(u,z) = \exp(-(d_u^2 + d_z^2) / (2 \cdot \text{bandwidth}^2))$ is
bounded above, on a slice at fixed $z$, by
$\exp(-d_z^2 / (2 \cdot \text{bandwidth}^2))$ regardless of $u$. At
`bandwidth = 0.08` and `z = 0.1`, that ceiling is only $\approx 0.46$,
and the weight averages closer to $0.15$ across the plotted range — so
on this particular slice the parametric term is never more than a
minority contributor. The calibrated bandwidth concentrates the
splice's action within roughly $0.1$–$0.15$ of the *exact* corner
point, and `z = 0.1` isn't close enough to it for the splice to move
the curve much. This is worth checking directly, at a slice that
actually is close enough:

```{r evt-splice-zoomed-corner, fig.height=8}
evt_zoom_df <- evt_results$evt_zoom_df

ggplot(evt_zoom_df, aes(u, density, color = source, linetype = source)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~edge, scales = "free", ncol = 2) +
  scale_color_manual(values = c("true" = "black", "hdcd (unspliced)" = "steelblue",
                                 "hdcd (evt-spliced)" = "firebrick")) +
  theme_minimal(base_size = 9) +
  theme(legend.position = "top") +
  labs(title = "Zoomed into the actual corner (z within 0.02 of 0 or 1)",
       x = expression(u[child]), y = "density")
```

**This is where you should be looking, and it reveals a real effect —
but not an improvement.** Right next to the true corner, the splice
visibly moves the curve on every edge. But look at *how*: on the
Clayton edges (left column), the evt-spliced curve actually dips
*below* the unspliced fit immediately around the true peak, then rises
to *overshoot* both the unspliced fit and the true density further out
along the shoulder — the mirror image on the Gumbel edges (right
column). A closer look at edge 1->2's exact fitted values makes this
precise:

```{r evt-splice-corner-numbers}
kable(evt_results$evt_corner_numbers,
      caption = "Edge 1->2 (Clayton), conditional density at z=0.02, near the corner")
```

Right at the peak (`u` around 0.03) the evt-spliced density is *lower*
than the unspliced fit, not higher — the opposite of what a
corner-strengthening splice is supposed to do — before crossing over
around `u = 0.15`-`0.17` and overshooting the true density by roughly
two orders of magnitude once `u` moves into the shoulder (true density
is near 0 there; the splice pushes the fit up to several times the
unspliced value). This is a genuine shape distortion, not simply "too
small an effect."

**A plausible explanation, not yet fully confirmed: Archimedean
densities are highly sensitive to theta specifically *off* the u = z
diagonal, deep in the corner.** The blend near the true corner is
almost entirely the parametric term (weight $\approx 0.97$ at
`u=0.005, z=0.02`), so its shape there is essentially the fitted
Clayton/Gumbel density's own shape, at the MLE-estimated theta (2.93
vs. a true 3.0 on this edge) rather than the true theta. A
theta estimate that looks close in absolute terms can still produce a
meaningfully different conditional-density *profile* well off the
diagonal (`u` far from `z`), where Archimedean families are most
sensitive to theta — plausibly explaining both the peak undershoot
(a lower theta concentrates less mass exactly at the corner) and the
shoulder overshoot (a lower theta spreads more mass into the region
just past it) simultaneously. This is offered as a hypothesis, not a
proven mechanism — it has not been isolated by, for instance, re-running
the splice with the true theta fixed instead of the MLE estimate.

**Bottom line, updated after actually looking at the right place:**
the splice is not simply "too weak to matter," as the `z=0.1`/`0.9`
slice alone would suggest — it has a real, visible effect exactly where
it's designed to act. But that effect is a distortion of the
conditional density's shape (undershoot at the peak, overshoot in the
shoulder) rather than a step toward the true corner spike, and the
correlation/`Delta_KL` gains reported above are, apparently, driven more
by this shoulder overshoot happening to correlate loosely with the true
curve's overall decay than by any genuine reproduction of the corner
itself.

Of the four interventions attempted across this notebook —
`lambda_roughness_grid` (ruled out entirely), `bernstein_degree_grid`
(real shape gain, real likelihood cost), `corner_relief` (weak,
noisy, directional evidence only), and now a true parametric EVT
splice with the objectively correct family and an accurate-looking
fit — **none resolves the sharp Clayton/Gumbel corner under-fit** in a
way that would be visually convincing up close. The EVT splice's
failure mode is the most informative of the four: it shows that even
getting the family and the point-estimate of theta right is not enough
when the raw kernel blend is evaluated at fine resolution right where
tail dependence concentrates, because the composed fit's local shape
turns out to be far more sensitive to the residual MLE error than the
theta-recovery numbers alone would suggest. That is a more useful
finding for a paper about the genuine difficulties of high-dimensional,
heavy-tailed modeling than a positive result would have been — and it
is a *second*, independent reason (alongside the design-scope one at
the top of this section) that this was the right investigation to run
and also the right feature to remove afterward: `hdcd`'s nonparametric
design was never going to admit a parametric-family fix as a permanent
feature, and empirically, even setting that constraint aside, the fix
did not cleanly work.

# Summary

```{r summary}
cat(sprintf("Dimension: d = %d, n = %d\n", d, n))
cat(sprintf("Total hdcd pipeline time: %.2f seconds\n", sum(unlist(timings))))
cat(sprintf("Skeleton edges recovered (any direction): %d / %d\n", n_exact + n_reversed, nrow(true_edges)))
cat(sprintf("Exact-direction edges: %d, reversed: %d, missed: %d, spurious: %d\n",
            n_exact, n_reversed, n_missed, n_spurious))
cat(sprintf("Delta_KL(true DAG vs. recovered DAG): %.4f\n", delta_kl))
cat(sprintf("Mean fitted-vs-true conditional density correlation: %.3f\n",
            mean(agreement$mean_correlation)))
```

**Takeaways:**

- The ground-truth generator was independently verified correct (uniform
  margins, matching Kendall's $\tau$) before being used to validate
  anything else.
- `hdcd` recovered the correct dependency *skeleton* for essentially all
  edges, in well under a minute for a 10-dimensional, 9-edge, five-family
  mixed vine copula on 2000 observations.
- Where the recovered DAG deviated from the true generative structure (a
  single reversed edge), the held-out KL score confirms it is a
  near-equally-good *statistical* factorization — precisely what the
  library's own documentation warns you to expect, and precisely why the
  reference DAG is never presented as a causal claim (spec sections 19, 34).
- The fitted conditional copula densities closely track the true,
  closed-form densities on the smooth, no-tail-dependence edges (Frank,
  Gaussian: correlation $\geq 0.97$), and reasonably well overall
  (mean correlation $\approx 0.96$), confirming the centered-
  Bernstein/Sinkhorn machinery (Milestones 5-6) is doing its job.
- The clearest weak spot: on edges with strong Archimedean tail
  dependence (Clayton, Gumbel), the fitted density visibly under-shoots
  the true density's sharp corner spike. The first hypothesis — that the
  notebook's fixed, global `lambda_roughness = 0.15` was over-smoothing
  these edges specifically — did NOT hold up once actually tested with
  `hdcd`'s new per-node roughness-selection option: held-out
  log-likelihood, the honest criterion, confirms 0.15 is already close to
  optimal for these edges, and a lighter penalty measurably overfits.
  The remaining mismatch looks more like a basis-expressivity limit of a
  bounded degree-4 Bernstein polynomial than a tunable smoothing bug. See
  "Known limitation" above — including the correction to the original,
  less rigorous diagnosis.
- Built the requested follow-up (a non-global, LEARNED `lambda_roughness`
  option, spec section 18) directly in the C core with its own test
  coverage, not as a notebook-only hack — see `DECISIONS.md`. It works
  correctly (verified: per-node adaptivity, no leakage into the outer
  holdout, fully backward compatible when unused) and was the right tool
  to properly check the hypothesis above, even though the answer it gave
  was "no" rather than the expected "yes."
- Followed up by raising `bernstein_degree` itself (also built directly
  in the C core, tail-dependence-gated and searched jointly with
  `lambda_roughness`, per `DECISIONS.md`): this genuinely improves
  shape-agreement with the true density on exactly the flagged edges, but
  at the cost of *worse* held-out log-likelihood at this notebook's
  sample size — a real bias-variance tradeoff between reproducing the
  true corner's shape and generalizing to new data, not a fix. See
  "Following up" above for the concrete numbers and the two remaining
  paths (more data, or a proper EVT tail-splice) logged for later.
- Growing $n$ to resolve that tradeoff was itself inconclusive (one
  realization per $n$ reversed non-monotonically) — which motivated a
  cheaper, more surgical intervention: an anisotropic roughness penalty
  that relaxes smoothing specifically near the tail-dependence corner
  instead of adding coefficients everywhere. Properly replicated (6
  seeds x 4 sample sizes, plotted with honest $\pm 1$ SD bars — not the
  $\pm$ SE bars that would visually flatter it), `corner_relief=0.6`
  showed the highest mean shape-correlation at every $n$ tested with a
  held-out-likelihood cost that never exceeded 1 replicate SD — a
  consistent-direction-across-four-noisy-points result, not any single
  point clearing significance on its own, but still the first of the
  three interventions tried here to clear its own pre-registered bar.
  See "Does corner relief resolve the tradeoff, properly replicated?"
  above for why that distinction matters.
- Finally tried, and then removed, the one lever the first three
  interventions all avoided: an actual *parametric* copula family (a
  copula-level EVT tail-splice — genuine Clayton/Gumbel MLE fits
  blended into the raw Bernstein kernel near the tail-dependence
  corner, before Sinkhorn renormalizes). A first, miscalibrated attempt
  (`evt_splice_gate=0.1`, a `bandwidth` borrowed unjustified from
  `corner_relief`) both spliced onto edges with no genuine tail
  dependence and, more surprisingly, *hurt* shape-correlation on 3 of
  the 4 edges where it selected the objectively correct family with an
  accurate fitted parameter. Recalibrating (`gate=0.5`,
  `bandwidth=0.08`) fixed the false positives and produced a small,
  real, directionally-consistent correlation improvement with a
  favorable `Delta_KL` — but plotting the actual densities at the slice
  where the splice's calibrated bandwidth actually has weight (not the
  `z=0.1`/`0.9` slice used everywhere else in this notebook, which
  turned out to be too far from the corner to show the effect at all)
  revealed the "improvement" was a genuine shape distortion: the
  spliced fit *undershoots* the true density right at the peak, then
  *overshoots* it well past the peak, in both cases moving further from
  the truth, not closer — the correlation gain was apparently an
  artifact of the shoulder overshoot loosely tracking the true curve's
  decay. **The feature has since been removed from `hdcd` entirely,
  for a reason independent of this empirical failure: assuming a
  parametric copula family for the joint dependence structure is
  outside this library's nonparametric design** (spec section 3's EVT
  provision is for marginal CDFs only, not the copula density). This
  section's numbers now come from a saved run of
  `notebooks/evt_splice_experiment.R`, executed once while the feature
  still existed, rather than the (removed) live API — see "Does a true
  EVT tail-splice..." above and `DECISIONS.md` for the full story. Of
  all four interventions tried in this notebook, none resolves the
  sharp Clayton/Gumbel corner under-fit in a way that would look
  convincing up close, and the EVT splice's failure mode — correct
  family, accurate theta, still a shape distortion — is arguably the
  most representative finding of this whole notebook's investigation
  into the genuine difficulties of high-dimensional, heavy-tailed
  modeling.
