---
title: "Recovering a High-Dimensional Vine Copula with hdcd"
subtitle: "End-to-end validation: simulate a known mixed-family vine copula, recover it, and check the results"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.width = 9)
```

# Purpose

This notebook is an end-to-end validation exercise for the `hdcd` library
(this repository): we

1. simulate data from a **known, high-dimensional, mixed-family vine
   copula** (Clayton, Gumbel, Frank, Gaussian, and Student-*t* pair-copulas,
   with non-uniform, heterogeneous marginals on top),
2. **recover it** with `hdcd`'s full pipeline (marginal smoothing → copula
   transform → distance-correlation dependence matrix → persistent-topology
   ordering → simulated-annealing DAG search → fixed-DAG fit),
3. time every stage, and
4. check the result three ways: does the recovered DAG match the true
   dependency structure, does its fitted score compare well to fitting the
   *true* DAG directly, and do the fitted conditional copula densities
   actually match the known, closed-form true densities?

We report timings throughout and finish with conditional density plots
comparing hdcd's nonparametric fit against ground truth.

# Setup

```{r libraries}
# This notebook lives at <repo_root>/notebooks/vine_copula_recovery.Rmd.
this_file <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NA_character_)
if (is.na(this_file) || !nzchar(this_file)) this_file <- file.path(getwd(), "vine_copula_recovery.Rmd")
repo_root <- normalizePath(file.path(dirname(this_file), ".."), mustWork = FALSE)
if (!file.exists(file.path(repo_root, "r", "DESCRIPTION"))) {
  # Fallback: assume the working directory is already the repo root.
  repo_root <- normalizePath(getwd(), mustWork = FALSE)
}
stopifnot(file.exists(file.path(repo_root, "r", "DESCRIPTION")))

if (!requireNamespace("hdcd", quietly = TRUE)) {
  message("Installing hdcd from source (", file.path(repo_root, "r"), ") ...")
  install.packages(file.path(repo_root, "r"), repos = NULL, type = "source")
}

suppressMessages({
  library(hdcd)
  library(copula)
  library(ggplot2)
  library(reshape2)
  library(knitr)
})

set.seed(20260819)
```

# A known, high-dimensional, mixed-family vine copula

## Construction

We build the ground truth as a **first-order (D-vine, tree-1) Markov-chain
copula**: variables $1, \ldots, d$ in a fixed order, where $U_{i+1}$ is
generated from a bivariate pair-copula conditional on $U_i$ alone. This is
a genuine, valid vine copula (a truncated D-vine) — the simplest member of
the D-vine family — and by mixing *different* pair-copula families and
strengths across the chain we still get a rich, high-dimensional,
non-Gaussian dependency structure.

**Why not a full multi-tree R-vine?** We tried one first. Chaining
`copula::cCopula()`'s conditional-inverse (h-function) calls through a
second tree hit a genuine numerical fragility: `uniroot()` inside the
package's Archimedean inverse-Rosenblatt implementation would occasionally
fail (`f.lower = f(lower) is NA`) once an intermediate conditional value
landed very close to the $[0,1]$ boundary after two chained inversions.
Rather than ship a demo whose *ground truth* might be silently wrong, we
kept to the tree-1 construction, which we verify below is exactly correct
(uniform marginals, matching Kendall's $\tau$), and mix five different
copula families across it for genuine complexity. Even a single-tree chain
directly exercises the part of `hdcd` we most want to validate: sparse DAG
recovery plus nonparametric conditional-copula shape estimation.

```{r vine-construction}
d <- 10L
n <- 2000L
eps <- 1e-6 # keeps chained h-function inversions off the exact [0,1] boundary

# One pair-copula per edge (i, i+1), i = 1..d-1: a deliberately varied mix.
families <- list(
  claytonCopula(3.0),   # 1 -> 2   strong lower-tail dependence
  gumbelCopula(2.2),    # 2 -> 3   upper-tail dependence
  frankCopula(5.0),     # 3 -> 4   symmetric, no tail dependence
  normalCopula(0.65),   # 4 -> 5   Gaussian
  tCopula(0.5, df = 4), # 5 -> 6   symmetric tail dependence (Student-t)
  claytonCopula(1.8),   # 6 -> 7   weaker lower-tail dependence
  gumbelCopula(3.0),    # 7 -> 8   strong upper-tail dependence
  frankCopula(3.5),     # 8 -> 9   symmetric, weaker
  normalCopula(0.45)    # 9 -> 10  Gaussian, weaker
)
family_names <- c("Clayton(3.0)", "Gumbel(2.2)", "Frank(5.0)", "Gaussian(0.65)",
                   "t(0.5, df=4)", "Clayton(1.8)", "Gumbel(3.0)", "Frank(3.5)", "Gaussian(0.45)")
stopifnot(length(families) == d - 1)

hinv <- function(u_given, w, cop) cCopula(cbind(u_given, w), copula = cop, indices = 2, inverse = TRUE)

w <- matrix(runif(n * d), ncol = d)
w <- pmin(pmax(w, eps), 1 - eps)

U <- matrix(NA_real_, n, d)
U[, 1] <- w[, 1]
for (i in 2:d) {
  u_prev <- pmin(pmax(U[, i - 1], eps), 1 - eps)
  U[, i] <- hinv(u_prev, w[, i], families[[i - 1]])
  U[, i] <- pmin(pmax(U[, i], eps), 1 - eps)
}
```

## Sanity check: is the ground truth actually correct?

A valid copula sample must have exactly Uniform(0,1) margins, and the
Kendall's $\tau$ between adjacent variables should match each pair-copula's
known $\tau$. If either check failed, nothing downstream would be worth
trusting.

```{r vine-sanity-check}
ks_pvalues <- sapply(1:d, function(j) ks.test(U[, j], "punif")$p.value)
empirical_tau <- sapply(1:(d - 1), function(i) cor(U[, i], U[, i + 1], method = "kendall"))
true_tau <- sapply(families, tau)

sanity <- data.frame(
  edge = paste0(1:(d - 1), " -> ", 2:d),
  family = family_names,
  true_kendall_tau = round(true_tau, 3),
  empirical_kendall_tau = round(empirical_tau, 3),
  marginal_KS_pvalue_ui = round(ks_pvalues[-d], 3)
)
kable(sanity, caption = "Ground-truth construction check: empirical vs. true Kendall's tau, per edge")
```

All Kolmogorov–Smirnov p-values should be comfortably large (no evidence
against uniformity), and the empirical Kendall's $\tau$ values should sit
close to their true values. This is our guarantee that the "true" copula
we're about to try to recover is exactly what we think it is.

## From copula scale to realistic, non-uniform data

`hdcd` is meant to work on **raw data**, not pre-transformed copula
coordinates — it fits its own marginals. So we push $U$ through a
heterogeneous set of marginal distributions (Normal, Exponential, and
several Normals with different means) to get a realistic-looking raw
dataset $X$, throwing away direct access to $U$ except for validation.

```{r apply-marginals}
X <- U
X[, 1] <- qnorm(U[, 1], mean = 0, sd = 2)
X[, 2] <- qexp(U[, 2], rate = 1)
for (j in 3:d) X[, j] <- qnorm(U[, j], mean = j, sd = 1)

colnames(X) <- paste0("X", 1:d)
kable(t(apply(X, 2, function(col) c(min = min(col), mean = mean(col), max = max(col)))),
      caption = "Raw data X: per-column summary (heterogeneous marginals)")
```

## What the pairwise dependence actually looks like

A quick look at three adjacent pairs, chosen to show three different
pair-copula shapes: Clayton (lower-tail clustering), Gumbel (upper-tail
clustering), and Gaussian (symmetric, elliptical).

```{r pairwise-scatter, fig.height=3.2}
pairs_df <- rbind(
  data.frame(u = U[, 1], v = U[, 2], edge = "1->2  Clayton(3.0)"),
  data.frame(u = U[, 2], v = U[, 3], edge = "2->3  Gumbel(2.2)"),
  data.frame(u = U[, 4], v = U[, 5], edge = "4->5  Gaussian(0.65)")
)
ggplot(pairs_df, aes(u, v)) +
  geom_point(alpha = 0.25, size = 0.6) +
  facet_wrap(~edge) +
  coord_equal() +
  theme_minimal() +
  labs(title = "Copula-scale pairwise dependence for three representative edges",
       x = expression(U[i]), y = expression(U[i+1]))
```

# Fitting hdcd

We now forget we ever saw `U` or the true DAG, and hand `hdcd` only the raw
data `X`. We run the pipeline stage by stage (rather than calling the
single `hdcd_fit()` wrapper) purely so we can time each stage separately;
`hdcd_fit()` performs exactly the same steps internally.

```{r fit-stages}
timings <- list()
tic <- function() Sys.time()
toc <- function(t0, label) {
  elapsed <- as.numeric(difftime(Sys.time(), t0, units = "secs"))
  timings[[label]] <<- elapsed
  invisible(elapsed)
}

# ---- Stage 1: marginal fitting (spec section 2) ----
t0 <- tic()
marginals <- lapply(1:d, function(j) hdcd:::.new_marginal(X[, j]))
toc(t0, "1. Marginal fitting (d columns)")

# ---- Stage 2: copula transform (spec section 4) ----
t0 <- tic()
U_hat <- matrix(NA_real_, n, d)
for (j in 1:d) U_hat[, j] <- hdcd:::.transform_to_copula(marginals[[j]], X[, j])
toc(t0, "2. Copula transform (d columns)")

# ---- Stage 3: dependence matrix (spec section 5) ----
t0 <- tic()
dm_ext <- hdcd:::.compute_dependence_matrix(U_hat, n, d)
toc(t0, "3. Distance-correlation dependence matrix")

# ---- Stage 4: persistent-topology ordering (spec section 6) ----
t0 <- tic()
topo_ext <- hdcd:::.compute_topology(dm_ext)
ordering <- hdcd:::.topology_ordering(topo_ext)
toc(t0, "4. Persistent-topology ordering")

# ---- Stage 5: simulated-annealing DAG search (spec sections 7, 17) ----
max_parents <- 2L
lambda_edge <- 0.05
bernstein_degree <- 4L
lambda_roughness <- 0.15
holdout_fraction <- 0.25
seed <- 1L
local_seed <- seed + 1L

t0 <- tic()
annealed <- hdcd:::.run_annealing_c(
  U_hat, n, d, ordering, max_parents, lambda_edge,
  bernstein_degree, lambda_roughness, holdout_fraction, local_seed,
  initial_temperature = 0.5, cooling_rate = 0.95,
  max_iterations = 600L, restarts = 3L,
  p_add = 1.0, p_remove = 1.0, p_swap = 1.0, anneal_seed = seed
)
toc(t0, "5. Simulated-annealing DAG search (1800 total proposals)")

# ---- Stage 6: final fixed-DAG fit (spec sections 14, 28) ----
t0 <- tic()
dag_fit_ext <- hdcd:::.dag_fit_c(U_hat, n, d, annealed$dag, bernstein_degree,
                                  lambda_roughness, holdout_fraction, local_seed)
toc(t0, "6. Final fixed-DAG fit")

timing_df <- data.frame(
  stage = names(timings),
  seconds = round(unlist(timings), 3)
)
timing_df <- rbind(timing_df, data.frame(stage = "TOTAL", seconds = round(sum(unlist(timings)), 3)))
kable(timing_df, caption = sprintf("Stage-by-stage timing (d=%d, n=%d)", d, n))
```

```{r assemble-model}
# Assemble the same hdcd_model structure hdcd_fit() would return, so the
# rest of the notebook can use the normal public API.
model <- list(
  marginals = marginals, dag = annealed$dag, dag_fit = dag_fit_ext,
  dependence_matrix_ext = dm_ext, topology_ext = topo_ext, d = d,
  max_parents = max_parents, bernstein_degree = bernstein_degree,
  lambda_roughness = lambda_roughness, holdout_fraction = holdout_fraction,
  local_seed = local_seed, best_score = annealed$score,
  score_trace = annealed$score_trace, accepted_trace = annealed$accepted_trace,
  acceptance_rate = annealed$acceptance_rate, U = U_hat, X = X
)
class(model) <- "hdcd_model"
print(model)
```

# Did it recover the structure?

## Recovered vs. true DAG

```{r dag-comparison}
recovered_edges <- hdcd_dag(model)
true_edges <- cbind(1:(d - 1), 2:d)

edge_key <- function(e) paste(pmin(e[, 1], e[, 2]), pmax(e[, 1], e[, 2]), sep = "-")
recovered_skeleton <- edge_key(recovered_edges)
true_skeleton <- edge_key(true_edges)

comparison <- data.frame(
  true_edge = paste0(true_edges[, 1], " -> ", true_edges[, 2]),
  recovered_as = sapply(seq_len(nrow(true_edges)), function(i) {
    key <- true_skeleton[i]
    idx <- which(recovered_skeleton == key)
    if (length(idx) == 0) return("MISSED")
    re <- recovered_edges[idx, ]
    if (re[1] == true_edges[i, 1]) "exact match" else sprintf("reversed (%d -> %d)", re[1], re[2])
  })
)
kable(comparison, caption = "Every true edge: was it recovered, and in the correct direction?")

n_exact <- sum(comparison$recovered_as == "exact match")
n_reversed <- sum(grepl("reversed", comparison$recovered_as))
n_missed <- sum(comparison$recovered_as == "MISSED")
n_spurious <- nrow(recovered_edges) - (n_exact + n_reversed)

cat(sprintf(
  "Skeleton edges recovered: %d / %d (%d exact direction, %d reversed). Spurious extra edges: %d.\n",
  n_exact + n_reversed, nrow(true_edges), n_exact, n_reversed, n_spurious
))
```

A reversed edge is not a bug: `hdcd`'s reference DAG is an optimized
**statistical factorization for density estimation**, not a causal claim
(this is stated explicitly throughout the library's own documentation and
tested directly in Milestone 9). For a chain, the pairwise dependence
between $U_i$ and $U_{i+1}$ is symmetric regardless of which one is
labelled "parent," so a direction flip on one edge — while everything
else matches — is exactly the kind of statistically-near-equivalent
alternative factorization the annealing search is allowed to land on.

## Dependence matrix and persistent-topology ordering

```{r dependence-heatmap, fig.height=5}
dm <- hdcd_dependence_matrix(model)
dm_long <- melt(dm, varnames = c("row", "col"), value.name = "dCor")
ggplot(dm_long, aes(col, row, fill = dCor)) +
  geom_tile() +
  scale_fill_viridis_c(limits = c(0, 1)) +
  scale_y_reverse() +
  coord_equal() +
  theme_minimal() +
  labs(title = "Fitted pairwise distance-correlation matrix", x = "column", y = "column")
```

The chain structure is clearly visible as a bright band along the
super-/sub-diagonal (adjacent variables are strongly dependent), fading
away from it — exactly the pattern a first-order Markov chain should
produce.

```{r ordering}
cat("Persistent-topology ordering:", hdcd_ordering(model), "\n")
cat("(true generative order was 1, 2, ..., 10 -- any ordering consistent\n",
    " with the chain's actual dependency strengths is a valid recovery.)\n")
```

## Did the annealing search actually converge?

```{r annealing-trace, fig.height=3.5}
trace_df <- data.frame(iteration = seq_along(model$score_trace), score = model$score_trace)
ggplot(trace_df, aes(iteration, score)) +
  geom_line(alpha = 0.5) +
  geom_hline(yintercept = model$best_score, color = "firebrick", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Simulated-annealing score trace (J(G) after each iteration)",
       subtitle = sprintf("acceptance rate = %.2f; dashed line = best score found (%.4f)",
                           model$acceptance_rate, model$best_score),
       x = "iteration (across all restarts)", y = "J(G)  [lower is better]")
```

$J(G)$ should trend down and then flatten out as the search converges on a
good structure — a rising-then-flat noisy trace, not one that's still
improving steeply at the end, is what confirms the iteration budget was
sufficient.

# How good is the recovered DAG's fit, really?

Structural mismatches on individual edges don't necessarily mean a worse
*fit*. We can check directly: fit the true DAG explicitly with
`hdcd_fit_dag()` (over the exact same training data) and compare held-out
KL scores with `hdcd_score_dag()` (spec section 19).

```{r true-dag-fit}
t0 <- tic()
true_fit <- hdcd_fit_dag(model, true_edges)
true_fit_time <- toc(t0, "True-DAG fit_dag() (for comparison only)")

delta_kl <- hdcd_score_dag(model, true_fit)
cat(sprintf("fit_dag(true DAG) took %.2f seconds\n", true_fit_time))
cat(sprintf(
  "Delta_KL(true DAG vs. recovered/annealed DAG) = %.4f\n(negative means the TRUE dag scores BETTER; near 0 means the annealed DAG is nearly as good)\n",
  delta_kl
))
```

# Do the fitted conditional densities actually match?

This is the most direct test of `hdcd`'s core nonparametric machinery
(the centered Bernstein tensor kernel plus Sinkhorn normalization,
Milestones 5-6): for every true edge, does `hdcd`'s fitted conditional
copula density $\hat c(u \mid z)$ — evaluated from the **true-DAG fit**,
so this is purely about density-shape accuracy, independent of the
discrete structure-search question above — match the known, closed-form
conditional density of the true pair-copula?

```{r conditional-density-grid, fig.height=14}
u_grid <- seq(0.02, 0.98, length.out = 60)
q_values <- c(0.1, 0.5, 0.9)

cond_rows <- list()
for (i in seq_len(d - 1)) {
  node <- i + 1L # child column (1-indexed), matches true_fit's node numbering
  for (q in q_values) {
    fitted <- exp(hdcd:::.local_fit_conditional_log_density(true_fit$dag_fit, node, u_grid, q))
    true_dens <- dCopula(cbind(rep(q, length(u_grid)), u_grid), families[[i]])
    cond_rows[[length(cond_rows) + 1]] <- data.frame(
      edge = sprintf("%d: %d->%d  %s", i, i, i + 1, family_names[i]),
      q = factor(sprintf("z = %.1f", q), levels = sprintf("z = %.1f", q_values)),
      u = u_grid,
      density = c(fitted, true_dens),
      source = rep(c("hdcd (fitted)", "true"), each = length(u_grid))
    )
  }
}
cond_df <- do.call(rbind, cond_rows)

ggplot(cond_df, aes(u, density, color = source, linetype = source)) +
  geom_line(linewidth = 0.7) +
  facet_grid(edge ~ q, scales = "free_y") +
  scale_color_manual(values = c("hdcd (fitted)" = "steelblue", "true" = "black")) +
  theme_minimal(base_size = 9) +
  theme(strip.text.y = element_text(angle = 0, hjust = 0), legend.position = "top") +
  labs(title = "Fitted vs. true conditional copula density c(u | z) for every true edge",
       x = expression(u[child]), y = "density")
```

## Summary agreement metrics

```{r conditional-density-summary}
agreement <- do.call(rbind, lapply(seq_len(d - 1), function(i) {
  node <- i + 1L
  per_q <- sapply(q_values, function(q) {
    fitted <- exp(hdcd:::.local_fit_conditional_log_density(true_fit$dag_fit, node, u_grid, q))
    true_dens <- dCopula(cbind(rep(q, length(u_grid)), u_grid), families[[i]])
    c(cor = cor(fitted, true_dens), rmse = sqrt(mean((fitted - true_dens)^2)))
  })
  data.frame(
    edge = sprintf("%d->%d", i, i + 1), family = family_names[i],
    mean_correlation = round(mean(per_q["cor", ]), 3),
    mean_rmse = round(mean(per_q["rmse", ]), 3)
  )
}))
kable(agreement, caption = "Per-edge agreement between hdcd's fitted density and the true density, averaged over z in {0.1, 0.5, 0.9}")

cat(sprintf("\nOverall mean correlation across all edges/conditioning values: %.3f\n",
            mean(agreement$mean_correlation)))
```

The per-edge table above is not uniform: the two smooth, no-corner-singularity
families (Frank, Gaussian) sit at correlation $\geq 0.97$, while the three
edges built from **tail-dependent** families (Clayton at edges 1 and 6,
Gumbel at edges 2 and 7) sit noticeably lower (0.90-0.96). That is not
random scatter — it is a real, systematic shape mismatch, worth
understanding rather than averaging away into the single headline number
above.

## Known limitation: sharp tail-dependence corners are under-fit

Look closely at the row-1 and row-6 panels (Clayton) at $z=0.1$, and the
row-7 panel (Gumbel) at $z=0.9$, in the density grid above: the **true**
density has a sharp, narrow spike right at the copula corner (Clayton's
lower-tail dependence concentrates mass as $u, z \to 0$; Gumbel's
upper-tail dependence does the same as $u, z \to 1$), while `hdcd`'s
**fitted** curve is visibly flatter and shorter there, even though it
tracks the true curve well everywhere else. This is the actual thing that
prompted a closer look at "densities that don't line up."

The first hypothesis was that a single, hand-picked, *global*
`lambda_roughness = 0.15` — applied identically to all nine edges — was
over-smoothing specifically these sharply-peaked edges, and that a
lighter, per-edge penalty would fit them better. `hdcd` now supports
exactly that as an option (`lambda_roughness_grid`, see `DECISIONS.md`):
a small validation grid, searched independently per node via an *inner*
train/validation split that never touches the outer holdout, entirely
outside the annealing search (spec section 18). Applying it directly
tests the hypothesis rather than just asserting it.

```{r roughness-grid-fit}
roughness_grid <- c(0.02, 0.05, 0.1, 0.15, 0.3, 0.6, 1.0)
t0 <- tic()
auto_fit <- hdcd:::.dag_fit_c(model$U, n, d, true_fit$dag, bernstein_degree, lambda_roughness,
                               holdout_fraction, local_seed,
                               lambda_roughness_grid = roughness_grid,
                               roughness_validation_fraction = 0.3)
auto_fit_time <- toc(t0, "Per-node auto lambda_roughness fit_dag() (diagnostic only)")

selected <- do.call(rbind, lapply(seq_len(d - 1), function(i) {
  node <- i + 1L
  data.frame(
    edge = sprintf("%d->%d", i, i + 1), family = family_names[i],
    selected_lambda = hdcd:::.local_fit_selected_lambda_roughness(auto_fit, node),
    holdout_score_auto = round(hdcd:::.dag_fit_holdout_scores(auto_fit, d)[node], 4),
    holdout_score_fixed_0.15 = round(hdcd:::.dag_fit_holdout_scores(true_fit$dag_fit, d)[node], 4)
  )
}))
kable(selected, caption = sprintf(
  "Per-node lambda_roughness selected from {%s} by inner-validation, vs. this notebook's fixed 0.15",
  paste(roughness_grid, collapse = ", ")
))
cat(sprintf("Delta_KL(fixed@0.15 vs. auto-selected) = %.4f\n",
            hdcd:::.dag_fit_kl_difference(true_fit$dag_fit, auto_fit)))
```

**The hypothesis does not survive contact with the data.** Every edge —
Clayton and Gumbel included — selects a lambda at or above this
notebook's original 0.15, not below it; several land higher (0.3). A
finer, continuous sweep (fixed `lambda_roughness`, no grid search)
confirms this is a genuine interior optimum, not an artifact of the
grid's spacing:

```{r roughness-fine-sweep, fig.height=3.2}
fine_lambdas <- c(0.01, 0.02, 0.05, 0.1, 0.15, 0.3, 0.6, 1.0, 2.0)
fine_sweep <- do.call(rbind, lapply(c(edge1 = 2L, edge7 = 8L, edge3 = 4L), function(node) {
  do.call(rbind, lapply(fine_lambdas, function(lr) {
    fit <- hdcd:::.dag_fit_c(model$U, n, d, true_fit$dag, bernstein_degree, lr, holdout_fraction, local_seed)
    data.frame(edge = c(`2` = "1->2 Clayton(3.0)", `8` = "7->8 Gumbel(3.0)", `4` = "3->4 Frank(5.0)")[[as.character(node)]],
               lambda_roughness = lr,
               holdout_score = hdcd:::.dag_fit_holdout_scores(fit, d)[node])
  }))
}))
ggplot(fine_sweep, aes(lambda_roughness, holdout_score)) +
  geom_line() + geom_point() +
  geom_vline(xintercept = 0.15, linetype = "dashed", color = "firebrick") +
  scale_x_log10() +
  facet_wrap(~edge, scales = "free_y") +
  theme_minimal() +
  labs(title = "Held-out score vs. lambda_roughness (log scale), fixed bernstein_degree",
       subtitle = "dashed line = this notebook's fixed 0.15",
       x = "lambda_roughness (log scale)", y = "held-out log-density score")
```

The same interior-optimum shape shows up for the Clayton edge, the
Gumbel edge, and even the smooth Frank edge (which has no corner at
all) — none of them want much lighter regularization than 0.15, and all
of them clearly punish going much lighter. `Delta_KL` in the table above
is slightly *positive* for the auto-selected fit relative to the fixed
default: per-node selection pays a real cost here (each candidate is fit
on a smaller *inner*-train subset, which is why it exists to begin with
— see `DECISIONS.md`), and on this dataset that cost isn't repaid,
because 0.15 was already close to optimal everywhere it was tested.

**Revised conclusion.** The original diagnosis — eyeballing the fitted
curve against the true curve and concluding "a lighter penalty would fit
this corner better" — was measuring agreement with a density that is only
knowable here because this is a synthetic exercise with ground truth.
*Held-out log-likelihood*, the criterion `hdcd` (and any real, non-
synthetic application) actually optimizes, disagrees: it says 0.15 is
already a good choice for these edges, lighter regularization measurably
overfits, and the visual corner-flattening is not primarily a smoothing
problem. What's left is more consistent with a basis-expressivity
limit — a degree-4 centered Bernstein tensor is a smooth, globally
bounded polynomial surface, and no amount of retuning `lambda_roughness`
lets a bounded polynomial reproduce an unbounded corner singularity. This
is a more honest, if less tidy, takeaway than the first pass reached, and
exactly what building the feature properly (rather than assuming the fix)
was for.

# Summary

```{r summary}
cat(sprintf("Dimension: d = %d, n = %d\n", d, n))
cat(sprintf("Total hdcd pipeline time: %.2f seconds\n", sum(unlist(timings))))
cat(sprintf("Skeleton edges recovered (any direction): %d / %d\n", n_exact + n_reversed, nrow(true_edges)))
cat(sprintf("Exact-direction edges: %d, reversed: %d, missed: %d, spurious: %d\n",
            n_exact, n_reversed, n_missed, n_spurious))
cat(sprintf("Delta_KL(true DAG vs. recovered DAG): %.4f\n", delta_kl))
cat(sprintf("Mean fitted-vs-true conditional density correlation: %.3f\n",
            mean(agreement$mean_correlation)))
```

**Takeaways:**

- The ground-truth generator was independently verified correct (uniform
  margins, matching Kendall's $\tau$) before being used to validate
  anything else.
- `hdcd` recovered the correct dependency *skeleton* for essentially all
  edges, in well under a minute for a 10-dimensional, 9-edge, five-family
  mixed vine copula on 2000 observations.
- Where the recovered DAG deviated from the true generative structure (a
  single reversed edge), the held-out KL score confirms it is a
  near-equally-good *statistical* factorization — precisely what the
  library's own documentation warns you to expect, and precisely why the
  reference DAG is never presented as a causal claim (spec sections 19, 34).
- The fitted conditional copula densities closely track the true,
  closed-form densities on the smooth, no-tail-dependence edges (Frank,
  Gaussian: correlation $\geq 0.97$), and reasonably well overall
  (mean correlation $\approx 0.96$), confirming the centered-
  Bernstein/Sinkhorn machinery (Milestones 5-6) is doing its job.
- The clearest weak spot: on edges with strong Archimedean tail
  dependence (Clayton, Gumbel), the fitted density visibly under-shoots
  the true density's sharp corner spike. The first hypothesis — that the
  notebook's fixed, global `lambda_roughness = 0.15` was over-smoothing
  these edges specifically — did NOT hold up once actually tested with
  `hdcd`'s new per-node roughness-selection option: held-out
  log-likelihood, the honest criterion, confirms 0.15 is already close to
  optimal for these edges, and a lighter penalty measurably overfits.
  The remaining mismatch looks more like a basis-expressivity limit of a
  bounded degree-4 Bernstein polynomial than a tunable smoothing bug. See
  "Known limitation" above — including the correction to the original,
  less rigorous diagnosis.
- Built the requested follow-up (a non-global, LEARNED `lambda_roughness`
  option, spec section 18) directly in the C core with its own test
  coverage, not as a notebook-only hack — see `DECISIONS.md`. It works
  correctly (verified: per-node adaptivity, no leakage into the outer
  holdout, fully backward compatible when unused) and was the right tool
  to properly check the hypothesis above, even though the answer it gave
  was "no" rather than the expected "yes."
